\chapter{Evaluation}
\label{ch:evaluation}

This chapter evaluated whether MERIT actually addresses the three gaps that were pointed out
earlier in Section~\ref{sec:relevance-to-merit}. 
Each of MERIT's pillars maps to a research question (RQ), which is derived from the literature
review earlier. Our studies are presented bottom-up (Ingestion $\rightarrow$ Fusion $\rightarrow$ 
Explainability $\rightarrow$ Human Alignment $\rightarrow$ Fairness and Scalability). Note that
this may disagree with the numerical prefixes of the studies in the repository, which follow the 
implementation order instead. Graphical representations of the studies are deferred to the appendix, 
and also cited inline.

\subsection{Evaluation Design and Research Questions}
\label{subsec:eval-design-rqs}

\textbf{RQ1 (Extensibility):} Can JD-driven, configurable metrics and reliable document ingestion
aid in the scoring of candidates? Gugnani, Misra and Lo et al.~\cite{gugnani2020,lo2025aihiring} criticise
the rigidity of scoring, as well as its constraints to fixed-dimension scoring. Resume parsing benchmarks,
such as ResumeBench~\cite{emnlp2025resumebench} also document layout-driven extraction failures that
can bound any downstream metric system.

\textbf{RQ2 (Multi-source):} Does evidence-aware multi-source fusion improve ranking over
CV-only baselines when public work samples support or contradict self-reported claims? Here, we extend
the work of Montandon et al.~\cite{montandon2021_roles} by considering other signals, specifically
GitHub to address unreliable claims present on a candidate's CV~\cite{henle2019_resume_fraud,dunning2004}.

\textbf{RQ3 (Transparency):} Can glass-box explainability and transparency aid decision calibration relative
to that of a black-box system? We aim to address the transparency deficits that are documented in 
modern algorithmic hiring reviews~\cite{bogen2018,raghavan2020}. We also apply the coalition Shapley
formalism from explainable ML~\cite{lundberg2017unified} at the data-source level (Section~\ref{sec:explainable-ai-shapley}).

We map each RQ to prior work along with the associated studies and primary metrics to answer these RQs (Table~\ref{tab:eval-literature-traceability}).
We also report criteria that we use to determine the success of each RQ. Note that our success criteria are not
derived from literature, and are instead internal targets, since many results are directional due to the small sample size
in our evaluation. Both the Traditional ATS and Modern AI ATS baselines are simplified in-repository proxies
for the keyword-matching and AI-assisted ATS categories mentioned earlier in Section~\ref{sec:automated-screening}.
Expert-consensus rankings follow from Lo et al.'s~\cite{lo2025aihiring} evaluation paradigm of comparing automated
scores with HR judgement. High Spearman $\rho$ measures the agreement with that panel, but cannot be 
generalised due to the sample size.

\begin{table}[htbp]
    \centering
    \caption[Literature-to-evaluation traceability]{Literature-to-evaluation traceability: research questions, 
    prior work, studies, and success criteria. Study~08 tests temporal skill decay~\cite{arthur1998}, grouped 
    under RQ2 as an integrity sub-test. Study~15 tests UI proxy debiasing~\cite{highhouse2008stubborn}, not 
    Shapley attributions. Studies~02,~03, and~06 are engineering or adversarial supplements 
    (Section~\ref{sec:fairness-computational-rigour}).}
    \label{tab:eval-literature-traceability}
    \footnotesize
    \renewcommand{\arraystretch}{1.45}
    \setlength{\tabcolsep}{6pt}
    \setlength{\extrarowheight}{2pt}
    \def\tabularxcolumn#1{m{#1}}%
    \begin{tabularx}{\textwidth}{@{}%
        >{\centering\arraybackslash}m{0.8cm}
        >{\centering\arraybackslash\hsize=1.05\hsize}X
        >{\centering\arraybackslash}m{1.9cm}
        >{\centering\arraybackslash\hsize=1.0\hsize}X
        >{\centering\arraybackslash\hsize=0.95\hsize}X @{}}
        \hline
        \textbf{RQ} & \textbf{Gap / prior work} & \textbf{Studies} & \textbf{Primary metrics} & \textbf{Success criterion} \\
        \hline
        RQ1 &
        Gugnani/Lo rigid weighting, ResumeBench parsing &
        04, 05, 13 &
        Skill $F_1$, name match, $\rho$ vs.\ recruiters &
        Parsing $\geq 80\%$, metric alignment exploratory \\
        RQ2 &
        Montandon supplementary GitHub, CV self-report bias &
        {\renewcommand{\arraystretch}{1}%
        \begin{tabular}[c]{@{}c@{}}08, 12, 01A/B, \\[-0.35ex] 07--10\end{tabular}} &
        Rank shift, Bayesian walkthrough, Spearman $\rho$ &
        $\rho \geq 0.7$ vs.\ experts (directional) \\
        RQ3 &
        Bogen/Raghavan transparency deficit, Lundberg Shapley &
        11, 14, 15 &
        Axiom checks, DRR, SUS, prestige gap &
        SUS $\geq 4$/5, blind gap compression \\
        \hline
    \end{tabularx}
\end{table}

Our evaluation takes inspiration from software-engineering experimentation~\cite{wohlin_threats}
through construct isolation, where we use synthetic JSON personas to isolate scoring logic from 
the parsing noise of the ingestion engine.
We also take inspiration from Lo et al.'s~\cite{lo2025aihiring} evaluation paradigm, where they 
anonymise real resumes with an HR panel to evaluate the agreement between automated scores and HR judgement.
MERIT follows a related approach, where we create synthetic profiles based on real candidate data,
and then evaluate the agreement between the automated scores and HR judgement.
Threats to validity, including this design trade-off, are discussed in 
Section~\ref{sec:threats-to-validity}.

\subsection{Codebase Traceability}
\label{subsec:eval-codebase-traceability}

Each study maintains a one-to-one mapping with a prefixed subdirectory under \texttt{evaluation/} 
in the project repository~\cite{merit_repo} (e.g.\ Study~04 $\leftrightarrow$ \texttt{04-ir\_parser\_test/}).
The sole exception is Study~01, which is split into two sub-studies (01A and 01B) due to the depth of the 
evaluation.

\input{evaluation/subsections/foundational_accuracy.tex}
\input{evaluation/subsections/evaluating_multi_source_fusion.tex}
\input{evaluation/subsections/evaluating_explainability.tex}
\input{evaluation/subsections/human_machine_alignment.tex}
\input{evaluation/subsections/fairness_computational_rigour.tex}
\input{evaluation/subsections/threats_to_validity.tex}
