\section{Evaluating Pillar 2: Multi-Source Fusion}
\label{sec:evaluating-multi-source-fusion}

% Helper macros for evaluation tables (defined at top to prevent duplicate definition/undefined command errors)
\newcommand{\splithead}[2]{\begin{tabular}[c]{@{}c@{}}\textbf{#1} \\ \textbf{#2}\end{tabular}}
\newcommand{\splitcand}[2]{\begin{tabular}[c]{@{}l@{}}#1 \\ #2\end{tabular}}
\newcommand{\splitdesc}[2]{\begin{tabular}[c]{@{}l@{}}#1 \\ #2\end{tabular}}
\newcommand{\splitcell}[2]{\begin{tabular}[c]{@{}c@{}}\##1 \\ (#2)\end{tabular}}

This section evaluates both the mathematical and empirical validity of MERIT's multi-source model.
To ensure a rigorous evaluation, we will proceed, once again, with a bottom-up evaluation. We will start with a study on the
algorithmic sensitivity with skill decay. From there, we will move on to evaluate Bayesian fusion, and how the model deals
with conflicting signals for a specific metric in a configuration. Then, we will proceed to an algorithmic ablation study;
here, we will run four scoring engines against the same candidate pool and JD: a \textbf{Traditional ATS} baseline (keyword-based CV screening),
a \textbf{Modern AI ATS} baseline (semantic CV screening), MERIT with only CV signals, and MERIT with all sources.
When referring to ``all'' sources, we specifically refer to the integration of GitHub and LinkedIn signals.
With the same engine setup and candidate pool, we will then proceed with a data ablation study, where we will point out the success and failure cases of the model.
The labels \textbf{Traditional ATS} and \textbf{Modern AI ATS} denote in-repository CV screening baselines that replicate scoring logic typical of rule-based and AI-assisted Applicant Tracking Systems (ATS); they are not proprietary commercial products. MERIT is evaluated as a separate multimodal screening system.

These studies assume that candidate CVs were successfully extracted without error. These evaluations run on emulated, 
synthetic candidate JSON profiles created to systematically test specific edge cases, bypassing the PDF extraction layer 
of MERIT (Stage 1). This allows us to isolate the multi-source Bayesian fusion and other scoring engines from any parsing noise.
By assuming perfect extraction, we establish a clean mathematical baseline so we can verify that the scoring logic, temporal decay
and conflict resolution can be demonstrated and evaluated as intended, rather than being subject to the varied errors of PDF parsing.
Validity implications are in Section~\ref{subsec:internal-validity}.

\subsection{Study 08: Algorithmic Sensitivity and Skill Decay}
\label{subsec:study-08-skill-decay}

All humans are prone to skill degradation over time. For a technical role, if you are not frequently working with a
specific technology, your expertise will naturally get "rusty" the longer you are removed from it. For candidates, we
consider the technical skill of a candidate $S(t)$ to be a function of time $t$ (in years) since the skill was
last actively practised.
\begin{equation}
    S(t) = S_0 \cdot e^{-\lambda t}
\end{equation}
where $S_0$ is the initial unweighted skill score extracted by the ingestion engine, and $\lambda$ is the
skill decay constant. In this implementation, $\lambda$ is set to $\approx 0.25$, corresponding
to a skill half-life of exactly $t_{1/2} = 2.77$ years, as described in Section~\ref{subsec:skill-decay-modelling}.

To evaluate two candidates to compare the effects of this model, both candidates have the exact same CV history,
with the only differing factor being the active time of usage for these skills (evaluated against a study baseline year of 2024):
\begin{itemize}
    \item \textbf{Candidate A (Active Pro):} Demonstrates active, continuous Go and Kubernetes usage from
          2022 to 2024 ($t = 0$).
    \item \textbf{Candidate B (Stale Pro):} Possesses identical technical depth in Go and Kubernetes,
          but has not actively worked with either technology since their last recorded project in 2018 ($t = 6$ years).
\end{itemize}

Applying the temporal decay formula to Candidate B's Go skill will yield the following:
\begin{equation}
    S(6) = S_0 \cdot e^{-0.25 \times 6} = S_0 \cdot e^{-1.5} \approx 0.223 \cdot S_0
\end{equation}

This means that Candidate B's Go skill score is only 22.3\% of what it would be if they were actively working with Go.
It is likely that within 6 years, the candidate has forgotten some syntax of the language,
but also the common libraries and frameworks used within it. If this candidate were to be selected for the role,
they would likely require some upskilling in order to get up to speed with the current best practices for
Go, and would experience a ramp-up period in the initial months.





\subsection{Study 12: Bayesian Conflict Resolution}
\label{subsec:study-12-bayesian-conflict-resolution}

In the previous section, we demonstrated that MERIT employs a Bayesian evidence fusion model with Beta
distribution conjugate priors \cite{gelman2013bayesian} to resolve conflicting signals across unstructured and structured data channels
without resorting to ad-hoc heuristics. Here, we will evaluate the empirical correctness of this model.

We initialise our model with the following priors:
\begin{equation}
    \alpha_0 = 0.1, \quad \beta_0 = 0.1
\end{equation}

To verify the correctness of this model, we shall run a step-by-step mathematical walkthrough for a candidate
\textbf{Felix Vance} for the extensible requirement of Java, as defined in Table~\ref{tab:bayesian-walkthrough}.

\begin{table}[htbp]
    \centering
    \caption[Study~12: Bayesian parameter shifts (Felix Vance)]{Bayesian Parameter Shifts for Felix Vance, demonstrating the effects of contradictory 
    evidence on the expected value and uncertainty. The convergence of these parameter updates are visualised in
        Figure~\ref{fig:bayesian-fusion-convergence} (Appendix~\ref{appendix:bayesian-fusion-charts}).}
    \label{tab:bayesian-walkthrough}
    \vspace{1.5ex}
    \begin{tabular}{l|c|c|c|c|c|c}
        \hline
        \splithead{Evidence Step}{($i$)} & \splithead{Strength}{($s_i$)} & \splithead{Confidence}{($c_i$)} & \splithead{Alpha}{($\alpha$)} & \splithead{Beta}{($\beta$)} & \splithead{Fused}{Score} & \splithead{Uncertainty}{($\sigma$)} \\[1.5ex] \hline
        0. Prior                         & ---                           & ---                             & 0.100                         & 0.100                       & 0.500                    & 0.456                               \\[1.5ex]
        1. CV (Self-Report)              & 1.000                         & 0.500                           & 0.605                         & 0.105                       & 0.852                    & 0.271                               \\[1.5ex]
        2. GitHub (Conflict)             & 0.000                         & 0.900                           & 0.606                         & 1.006                       & 0.376                    & 0.300                               \\[1.5ex]
        3. LinkedIn (Endorse)            & 0.800                         & 0.200                           & 0.774                         & 1.054                       & 0.423                    & 0.294                               \\[1.5ex]
        \hline
    \end{tabular}
\end{table}

This walkthrough demonstrates three core behaviours of the model.

\subsubsection{Strong Signal Handling (CV Assertion)}
Felix's CV explicitly states proficiencies in Java, earning a strong positive signal of ($s_1 = 1.0$) for the CV signal.
This signal has a moderate reliability of ($c_1 = 0.5$) as it is self-reported, and has not been verified by any other source.
In this first step, we observe that the model correctly interprets the strong positive signal of $1.0$ from the CV,
increasing the $\alpha$ parameter from $0.1$ to $0.605$, while $\beta$ updates to $0.105$.
As a result, the match score, once merged with the CV signal, jumps from $0.5$ to $0.852$, highlighting that a strong signal
is rewarded accordingly. It is also vital to note that the uncertainty also decreases from $0.456$ to $0.271$, due to the
inclusion of some, albeit weak, evidence.

\subsubsection{Conflicting Signal Handling (GitHub)}
This is where the beauty of the Bayesian evidence model comes into play. Felix's GitHub language history features no
evidence of them ever working with Java; however, there is a strong indication of them working with C++ (a language that is not on the Job Description). Here, there is
a strong contradictory signal between the CV and the GitHub profile. Felix has a ($s_2 = 0.0$) for Java, for a data source
with a strong reliability of ($c_2 = 0.9$). Felix has provided a signal that strongly disagrees with his claims on his CV.
Due to this, the model updates the $\beta$ parameter from $0.1$ to $1.006$, while $\alpha$ remains at $0.606$.
This leads to the match score decreasing from $0.852$ to $0.376$, highlighting that the model correctly penalises a
contradictory signal. Another key observation is that the uncertainty also increases from $0.271$ to $0.300$,
analysing that the model correctly captures the increased uncertainty in the presence of contradictory evidence.

It is important to note here that the penalisation applied due to Bayesian conflict resolution is not intended to be
significant, since it may be the case that Felix may not have publicly available evidence of their skills due to the
nature of their work. It could be likely that Felix is a software engineer, and the projects he has worked on are proprietary,
and therefore not publicly available.

\subsubsection{LinkedIn Signal Confirmation}

In step 3, Felix has a relatively strong signal from LinkedIn ($s_3 = 0.8$) with a relatively low confidence ($c_3 = 0.2$).
This step is less important than the previous steps, however, we will provide it to show the robustness of the model.
Here, the parameters update to $\alpha = 0.774, \beta = 1.054$, leading to a fused score of $0.423$ with an
uncertainty of $0.294$ due to there being more claim-based evidence of potential understanding of Java.

We provide a low confidence here because LinkedIn is considered a social media platform, where candidates may often
inflate their skills in order to appear more attractive to potential employers. Also, unlike the one-page restriction applied
to a CV, there are no restrictions applied to the number of skills a candidate can list on their LinkedIn profile.

This worked example mathematically demonstrates that MERIT's Bayesian evidence fusion model does not just apply a simple
weighted average, or ignore conflicts. Instead, it captures, tracks and scales the uncertainty in the presence of
highly contradictory signals from sources of high confidence.

\subsection{Study 01A: Algorithmic Ablation}
\label{subsec:study-01a-algorithmic-ablation}

To evaluate the benefits and drawbacks of MERIT's multi-source fusion model, we will compare it against three other baseline screening engines.
The first is \textbf{Traditional ATS}, an in-repository keyword-matching baseline representing CV screening logic typical of rule-based ATS products \cite{vanesch2019}; it scores candidates solely on the count of matching keywords in their CV.
The second is \textbf{Modern AI ATS}, an in-repository semantic baseline representing CV-only screening typical of AI-assisted ATS products; it computes cosine
similarities between CV embeddings and Job Description (JD) embeddings using a Sentence-BERT (SBERT) model \cite{reimers2019sentence}.
Neither baseline is a commercial ATS platform. We also run \textbf{MERIT (CV Only)} on CV data alone, and \textbf{MERIT (Full)}, which considers all sources of
evidence.

The complete ranking outcomes, as well as the normalised candidate scores across these four models are detailed in
Table~\ref{tab:algorithmic-ablation}.




\begin{table}[htbp]
    \centering
    \caption[Study~01A: Algorithmic benchmark rankings]{ Algorithmic Benchmarking and Candidate Rankings for 10 personas across four different scoring engines. $\delta\%$ represents
    the score gain/loss between the CV-Only and Full MERIT configurations, highlighting the effect of multi-source fusion on different personas. 
    Refer to Figure~\ref{fig:rank-displacement-successes} (Appendix~\ref{appendix:rank_displacement}) for the complete rank displacement trajectories.
    Note that scores have been normalised on a scale of 0 to 100\% where appropriate.}
    \label{tab:algorithmic-ablation}
    \vspace{1.5ex}
    \small
    \setlength{\tabcolsep}{2.5pt}
    \begin{tabularx}{\textwidth}{l|l|c|c|c|c|c}
        \hline
        \textbf{Candidate}        & \textbf{Profile Characterisation}           & \splithead{Trad. ATS}{{(\%)}} & \splithead{Modern AI}{{(\%)}} & \splithead{CV-Only}{{(\%)}} & \splithead{Full MERIT}{{(\%)}} & \splithead{Fusion Shift}{{($\Delta$\%)}} \\[2.5ex] \hline
        \splitcand{Alex}{Rivers}  & \splitdesc{Verifiable Elite}{Developer}     & \splitcell{5}{55.0}           & \splitcell{1}{86.4}           & \splitcell{1}{40.3}         & \splitcell{1}{78.0}            & +37.7                                    \\[3.5ex]
        \splitcand{Felix}{Vance}  & \splitdesc{Identity Mismatch}{(Java/C++)}   & \splitcell{3}{60.0}           & \splitcell{6}{60.0}           & \splitcell{2}{37.4}         & \splitcell{3}{44.7}            & +7.3                                     \\[3.5ex]
        \splitcand{Jordan}{Smith} & \splitdesc{Hidden Gem}{(1-Year Experience)} & \splitcell{10}{3.0}           & \splitcell{9}{20.0}           & \splitcell{10}{10.5}        & \splitcell{2}{64.3}            & \textbf{+53.8}                           \\[3.5ex]
        \splitcand{Buzz}{Ward}    & \splitdesc{Adversarial}{Keyword Stuffer}    & \splitcell{2}{64.0}           & \splitcell{5}{60.0}           & \splitcell{3}{35.5}         & \splitcell{6}{36.2}            & +0.7                                     \\[3.5ex]
        \splitcand{Ghost}{Gary}   & \splitdesc{Unverified}{Assertions Only}     & \splitcell{4}{60.0}           & \splitcell{7}{60.0}           & \splitcell{5}{29.4}         & \splitcell{8}{29.4}            & 0.0                                      \\[3.5ex]
        \splitcand{Sam}{Old}      & \splitdesc{Stale Professional}{(Inactive)}  & \splitcell{6}{55.0}           & \splitcell{8}{60.0}           & \splitcell{7}{23.0}         & \splitcell{5}{36.8}            & +13.8                                    \\[3.5ex]
        \splitcand{Vince}{Vault}  & \splitdesc{High-Tenure}{Core Developer}     & \splitcell{1}{70.0}           & \splitcell{4}{68.4}           & \splitcell{4}{33.2}         & \splitcell{9}{20.6}            & \textbf{-12.6}                           \\[3.5ex]
        \splitcand{Fiona}{Frost}  & \splitdesc{Evidence of Absence}{(React)}    & \splitcell{8}{40.0}           & \splitcell{3}{68.4}           & \splitcell{6}{26.6}         & \splitcell{7}{34.0}            & +7.4                                     \\[3.5ex]
        \splitcand{Kim}{Junior}   & \splitdesc{Entry-Level}{High Potential}     & \splitcell{9}{11.5}           & \splitcell{10}{19.2}          & \splitcell{9}{13.4}         & \splitcell{4}{38.9}            & +25.5                                    \\[3.5ex]
        \splitcand{Carl}{Corp}    & \splitdesc{Closed-Source}{Corporate Giant}  & \splitcell{7}{40.0}           & \splitcell{2}{69.0}           & \splitcell{8}{17.1}         & \splitcell{10}{17.9}           & +0.8                                     \\[3.5ex]
        \hline
    \end{tabularx}
\end{table}

The results demonstrate how the layering of scoring configurations systematically neutralises adversarial manipulation
and corrects for temporal or formatting biases. In the Traditional ATS engine (which relies entirely on raw CV keyword
frequency), rankings are dominated by high-tenure profiles and adversarial profiles such as Buzz Ward (\#2), who
successfully exploits keyword-density screening. Under the semantic Modern AI ATS, cosine similarity correctly
evaluates conceptual alignment (elevating Alex Rivers to \#1 and deprioritising Buzz Ward to \#5), yet remains highly
vulnerable to unverified assertions (incorrectly elevating Carl Corp to \#2 despite a total lack of verifiable
professional footprint).

For the CV-Only baseline, the introduction of the keyword-stuffing audit penalises Buzz Ward's stuffed CV, but also
highlights a critical structural limitation: without secondary evidence, screening models struggle to differentiate
between genuine high-density expertise and adversarial optimization, occasionally penalising valid CVs. By transitioning
to the Full MERIT engine, the introduction of Bayesian Evidence Factors (BEFs) resolves this tension—allowing the
system to verify assertions against empirical external signals and insulate candidates from parsing noise.

Crucially, while analysing ordinal rank displacements highlights the recruiter's sorting experience, evaluating the
cardinal score adjustments ($\Delta$) between the CV-Only baseline and the Full MERIT multi-source configuration reveals
the true mechanistic sensitivity of the Bayesian fusion engine. Ordinal rankings can obscure the magnitude and conviction
of the engine's adjustments. For instance, Sam Old's rank shift is modest (+2 rank displacement, moving from \#7 to \#5),
yet his cardinal score rises by $+13.8\%$, representing a high-conviction verification of his inactive competencies that was
previously penalised by single-source temporal decay. Similarly, Jordan Smith's transition shows an extraordinary cardinal
leap of $+53.8\%$ (from $10.5\%$ to $64.3\%$), mathematically capturing the breakthrough potential of a junior candidate whose
sparse resume is validated by elite open-source contributions.

Conversely, the fusion engine successfully limits or penalises unverified or adversarial claims: Buzz Ward's stuffed CV
receives a negligible cardinal shift of $+0.7\%$, while Vince Vault experiences a negative cardinal shift of $-12.6\%$
($33.2\% \to 20.6\%$), representing the heavy evidence conflict penalty applied under validation failures. The specific
data-source mechanics driving these individual shifts are detailed in Section~\ref{subsec:study-01b-data-source-ablation}.
Figure~\ref{fig:rank-displacement-successes} (Appendix~\ref{appendix:rank_displacement}) traces the highlighted cases discussed above.

\subsection{Study 01B: Multi-Source Data Ablation}
\label{subsec:study-01b-data-source-ablation}

This study will cover the critical trade-offs present in a system that doesn't just follow claimed evidence but also takes
into account demonstrated evidence. MERIT aims to increase the amount of information to take into account for candidate evaluation
by considering LinkedIn and Github alongside CVs.

\subsubsection{Systemic Success: The Discovery of Jordan Smith}

The most significant success of multi-source integration lies in Jordan Smith, a junior developer with only 1 year of
formal experience, whose sparse resume is automatically discarded by both traditional and modern semantic ATS (ranking
\#10 and \#9 respectively) due to low tenure.
What is not mentioned on Jordan's CV, however, is his extensive work that is publicly available on his GitHub account, demonstrating
significant technical expertise in both Java and TypeScript, two key technologies in the JD for this study.

Once we consider this candidate with MERIT's full capabilities, namely GitHub, Jordan jumps from \textbf{Rank \#10} to \textbf{Rank \#2}.
The job description that prioritises the use of both Java and TypeScript now takes into account, and values, the evidence that a candidate
has to show for themself, and in this case, Jordan has the most demonstrated evidence among the candidate batch. By shifting the
evaluation paradigm from unverified skills towards verified demonstratable evidence, we see that MERIT is able to prevent valid
candidates from being lost in the traditional "Resume Black Hole."

\subsubsection{Systemic Limitation: The "Proprietary Data Gap" for Carl Corp}
However, the introduction of multi-source evaluation does come with its own set of trade-offs.
The achilles heel of this approach is that merit is \textbf{heavily dependent} on the accessibility of a candidate's data.
In this case, we point out Carl Corp, a candidate who's data on GitHub is essentially empty, since most of his commits are towards
private repositories. As a result, he scores low for GitHub signals, and is unable to demonstrate his expertise in the programming languages
required by the JD. This results in a fall from Rank \#2 to Rank \#10 between the Modern AI ATS and Full MERIT evaluations.
This "Propriety Data Gap" is a critical limitation of MERIT that will significantly, and unfairly hinder candidates that are not able
to share their evidence of code, and will unfairly advantage candidates who have less experience but are able to share their code.
This represents a key trade-off in the design of MERIT. While we are now able to avoid some forms of bias due to the multi-source nature
of the system, we are now introducing a new form of bias based on the accessibility of a candidate's data. This raises the question of
whether the benefits of multi-source evaluation outweigh the costs of introducing this new form of bias.


\subsubsection{Integrity Resolution}
\label{subsubsec:identity-mismatch-resolution}

Now that MERIT considers multiple sources of data, it opens up the possibility of candidates trying to game the system
by including false sources of evidence. This will be referred to as \textbf{Identity Squatting}. We will evaluate the effectiveness
of MERIT's integrity resolution mechanisms in detecting and penalising such candidates.

Vince Vault is a candidate who claims a strong 10 years of experience as a Lead Software Architect on his CV, and presents himself as a
Java expert. This will provide an interesting case study into the effectiveness of MERIT's integrity resolution mechanisms.
Here, Vince boasts excessive skills, scoring exceptionally well in a Traditional ATS (\textbf{Rank \#1}) and Modern AI ATS
(\textbf{Rank \#4}). However, it could be the case that Vince is actually experienced in C\#, and upon hearing about Java and C\# being
similar programming languages, he decides to try and game the system by applying to this role as a "Java expert."

To prevent GitHub from identifying his proficiencies in C\# and not Java, Vince instead looks for an opportunity to find a candidate who
demonstrates proficiencies in Java, and instead provides a link to someone else's GitHub profile. This is a serious case of identity
squatting and will be used to test the robustness of MERIT's integrity resolution mechanisms. Due to the severe nature of this act of
fraud, MERIT's Integrity Layer applies a flat veto penalty, deducting exactly 20 percentage points ($0.20$ absolute score reduction) from
his score, regardless of the evidence presented by the other sources. The result of this absolute penalisation is highly significant:
Vince's CV-Only score is $33.2\%$, and although the squatted GitHub profile temporarily inflates his raw average to $40.6\%$, the application
of the $20\%$ veto drags his final score to $20.6\%$, dropping his rank from Rank \#4 to Rank \#9. This ensures that, on the backend side,
MERIT is successfully able to identify and penalise candidates who try to game the system.

It is important to note that this is a rare edge case, it is not very common for candidates to engage in this type of fraud.
However, it is important to consider this edge case in the design of MERIT, as it is a potential threat to the integrity
of the system.