\section{Evaluating Pillar 3: Glass-Box Explainability \& HCI}
\label{sec:evaluating-explainability}

This section evaluates the practical utility of MERIT's explainability pillar. It also covers the mathematical correctness
as well as the deterministic behaviour of the explainability engine to ensure transparency and trust in the screening system. As
explained in Section~\ref{sec:explainable-ai-shapley}, MERIT employs game theory, specifically Shapley values, to provide
source-level attribution to each metric, as well as each candidate. We verify that the axioms of game theory are satisfied,
with a methodological justification regarding the symmetry axiom. Following this, we describe a human-computer
interaction (HCI) user trial conducted to evaluate the usability and decision-making impact of MERIT's explainability layer on professional
recruiters.

\subsection{Study 11: Shapley Value Verification}
\label{subsec:study-11-shapley-verification}

To ensure our source-level attributions are mathematically source, we shall formally evaluate the implementation against three of
the four axioms of cooperative game theory. For our settings, we denote the candidate's final match score as the "winning" of the game,
and the implemented data sources (GitHub, LinkedIn, CV) as the "players" ($N = \{CV, GitHub, LinkedIn\}$)

Our Shapley values are verified using our cohort of 10 candidates. We use the same 10 candidates from the previous section.

\subsubsection{Game-Theoretic Axioms and Verification Methodology}
\begin{enumerate}
    \item \textbf{Axiom 1: Efficiency.} The sum of all source attributions ($\phi_i(v)$) must exactly equal the
          difference between the value of the grand coalition $v(N)$ (the full match score) and the empty coalition $v(\emptyset)$.
          \begin{equation}
              \sum_{i \in N} \phi_i(v) = v(N) - v(\emptyset)
          \end{equation}

          Since the candidate's empty coalition score is guaranteed to be zero by the system's zero-baseline rule ($v(\emptyset) = 0$),
          we must have that the attributions sum exactly to the candidate's final match score. That is, $\phi_{CV} +
              \phi_{GitHub} + \phi_{LinkedIn} = v(N)$.

    \item \textbf{Axiom 2: Symmetry.} If two sources $i$ and $j$ make identical marginal contributions to all
          possible sub-coalitions, they must receive identical attributions:
          \begin{equation}
              \text{If } v(S \cup \{i\}) = v(S \cup \{j\}) \quad \forall S \subseteq N \setminus \{i, j\}, \quad \text{then } \phi_i(v) = \phi_j(v)
          \end{equation}

          MERIT's players were designed to contribute to a candidate's final match score in distinct ways. Some data sources
          are leveraged as they are structured representations of a candidate's professional history (LinkedIn and CV).
          Other data sources are leveraged for their representations of a candidate's technical skills and expertise
          (GitHub). Because of this, two real sources will never have identical marginal contributions to all
          possible coalitions. Therefore, the symmetry axiom is not tested in this evaluation because our
          heterogeneous players (CV, GitHub, and LinkedIn) contribute to candidate scores in fundamentally different
          ways, making symmetric marginal contributions impossible.

    \item \textbf{Axiom 3: Null Player (Dummy Player).} A source that contributes zero marginal value to all possible
          coalitions must receive an attribution of exactly zero:
          \begin{equation}
              \text{If } v(S \cup \{i\}) = v(S) \quad \forall S \subseteq N \setminus \{i\}, \quad \text{then } \phi_i(v) = 0
          \end{equation}
          We verify this by analyzing candidates with completely blank or missing profiles on specific channels
          (e.g., Ghost Gary, who lacks any public GitHub or LinkedIn footprint), verifying that their attributions are exactly zero.

    \item \textbf{Axiom 4: Additivity (Per-Metric Efficiency).} If a game is represented as the sum of independent sub-games
          ($v + w$), the attributions must sum additively: $\phi_i(v + w) = \phi_i(v) + \phi_i(w)$.
          Because MERIT calculates overall candidate scores by aggregating multiple individual skill and role metrics, we
          verify this property at the granular level. For each active metric $m$, the sum of per-metric Shapley values across
          sources must equal the grand coalition score for that specific metric:
          $\sum_{i \in N} \phi_{m, i} = v_m(N)$.

    \item \textbf{Extended Computational Property: Determinism.} The attribution engine must produce identical Shapley
          values across repeated executions under identical input data. We test this by running the scoring and attribution
          routine 5 times for the entire cohort. Because $N=3$ data sources permits exact, analytical calculation rather than
          stochastic Monte Carlo sampling, the maximum observed drift is expected to be exactly zero.
\end{enumerate}


\subsubsection{Quantitative Verification Results}
After implementing the source attribution features, we ran our suite of quantitative tests to verify that the
attribution engine satisfied the core axioms of game theory. The quantitative results for Axioms 1, 3, and 4 are provided
in Table~\ref{tab:shapley-verification}.

\begin{table}[htbp]
    \centering
    \caption[Study~11: Shapley verification]{The sum of all source attributions ($\sum \phi_i$) matching the grand coalition match score
    $v(N)$ within floating-point tolerance ($\Delta < 10^{-6}$) to account for precision errors. Note that negative Shapley values are present,
    noting Vince Vault's penalisation for an identity mismatch on GitHub.}
    \label{tab:shapley-verification}
    \vspace{1.5ex}
    \small
    \setlength{\tabcolsep}{5pt}
    \begin{tabular}{l|c|c|c|c|c|c|c}
        \hline
        \textbf{Candidate Name} & $\phi_{CV}$ & $\phi_{GitHub}$ & $\phi_{LinkedIn}$ & $\Sigma \phi_i$ & $v(N)$ & $\Delta$             & \textbf{Status} \\ \hline
        Alex Rivers             & +0.4432     & +0.2952         & +0.0087           & 0.7470          & 0.7470 & $0.00 \times 10^{0}$ & PASS            \\
        Buzz Ward               & +0.4845     & 0.0000          & +0.0075           & 0.4920          & 0.4920 & $0.00 \times 10^{0}$ & PASS            \\
        Carl Corp               & +0.1635     & 0.0000          & +0.0155           & 0.1790          & 0.1790 & $0.00 \times 10^{0}$ & PASS            \\
        Felix Vance             & +0.4805     & +0.0660         & +0.0075           & 0.5540          & 0.5540 & $0.00 \times 10^{0}$ & PASS            \\
        Fiona Frost             & +0.2660     & +0.0660         & +0.0080           & 0.3400          & 0.3400 & $0.00 \times 10^{0}$ & PASS            \\
        Ghost Gary              & +0.4010     & 0.0000          & 0.0000            & 0.4010          & 0.4010 & $0.00 \times 10^{0}$ & PASS            \\
        Jordan Smith            & +0.1030     & +0.4095         & +0.0115           & 0.5240          & 0.5240 & $0.00 \times 10^{0}$ & PASS            \\
        Kim Junior              & +0.1880     & +0.0490         & 0.0000            & 0.2370          & 0.2370 & $0.00 \times 10^{0}$ & PASS            \\
        Sam Old                 & +0.1648     & +0.1878         & +0.0153           & 0.3680          & 0.3680 & $0.00 \times 10^{0}$ & PASS            \\
        Vince Vault             & +0.4257     & -0.0708         & +0.0112           & 0.3660          & 0.3660 & $0.00 \times 10^{0}$ & PASS            \\ \hline
    \end{tabular}
\end{table}

\begin{itemize}
    \item \textbf{Efficiency Verification:} For all 10 candidates, the maximum absolute delta $\Delta = |\sum \phi_i - v(N)|$
          was within floating-point tolerance ($< 10^{-6}$), confirming perfect efficiency. No phantom scores were present,
          therefore proving the mathematical validity of the engine's zero-baseline guarantee.

    \item \textbf{Null Player Verification:} Consider Ghost Gary, a candidate who has empty, but verified, LinkedIn and GitHub profiles.
          This results in $\phi_{LinkedIn} = 0.0000$ and $\phi_{GitHub} = 0.0000$. This result highlights that sources which provide no
          marginal evidence shall not receive any source attribution, therefore satisfying Axiom 3 (the Null Player axiom).

    \item \textbf{Determinism Verification:} Over 5 repeated executions, the maximum observed drift in any candidate's Shapley value
          was exactly $0.00 \times 10^0$. This confirms that the XAI engine is perfectly deterministic, which is mathematically 
          guaranteed since exact, analytical calculations are performed rather than stochastic Monte Carlo approximations.

    \item \textbf{Per-Metric Efficiency:} Per-metric Shapley sums matched the grand coalition metric scores with 100\% accuracy
          across all candidate-metric pairs (e.g., matching the exact Python or Go individual scores).
\end{itemize}

\subsubsection{Key XAI Insights: The Power of Negative Attribution}

The source-level attribution breakdown for each candidate is in Figure~\ref{fig:shapley-attribution-breakdown}
(Appendix~\ref{appendix:shapley-charts}), which illustrates how the Glass-Box 
scorecard surfaces drivers behind each score.

A key finding from this study was the presence of negative attributions within the candidate's source attributions. In the context
of cooperative game theory, a player would only receive a negative Shapley value if their inclusion in a coalition would lead to a
reduction in the overall coalition's winnings. In the context of MERIT, this would occur when the introduction of a source would lead
to contradictory information that would trigger a penalty, thus reducing the overall score.

To illustrate this, consider Vince Vault, a candidate who claims, and backs up elite Java expertise on their CV. However, they use
evidence from another candidate's GitHub profile to demonstrate this, stealing their work and passing it off as their own.
While Vince's CV-only score is relatively high with $\phi_{CV} = 0.426$, the introduction of malicious intent via squatting
triggers an identity mismatch penalty, leading to a drop in the overall score. Alongside this, the introduction of this
penalty is also passed on to the frontend to be visualised and investigated to ensure actual intent of malice, allowing the
recruiter to make an informed decision.

Another core finding is, like before, Jordan Smith. As shown in previous studies, Jordan Smith is a 'hidden gem' candidate,
usually scoring poorly in CV-only evaluation, however scoring highly when GitHub is included. The evidence of this can be
shown by his CV contribution of $\phi_{CV} = 0.103$, while his GitHub contribution is $\phi_{GitHub} = 0.410$. This enables
recruiters to instantly see that Jordan's high overall score is backed by demonstrated evidence, rather than just self-reported
resume claims.

It is worth pointing out that LinkedIn profiles for candidate personas were intentionally designed with highly sparse, or blank,
profile data. This was done to highlight the difference between claimed and demonstrated evidence, and to show that
MERIT can function effectively even with limited data. It also highlights that LinkedIn profiles should be used as a
secondary source of evidence for inaccurate parsing of unstructured data (CVs), rather than a primary source of evidence.
Conceptually, we wished to highlight the core architectural distinction between the role of claimed evidence (CVs and LinkedIn)
and demonstrated evidence (GitHub). Both CVs and LinkedIn profiles typically represent the same self-reported, claimed evidence,
and are authored by the candidate themself. They are subject to the same faking, exaggerations and keyword-stuffing behaviours.
As a result, LinkedIn data was often neglected in this study, since there is little need to enforce the same signals in LinkedIn
which would already be present in the synthetic data created for these personas. Contrasting to this, GitHub represents the
independent, demonstrated evidence, providing us with means to calculate the complexity of a candidate's code, their language
utilisation as well as project history. MERIT is designed to expose this which cannot be falsified, therefore pointing out
the importance of additional evidence sources to create a holistic picture of a candidate.

\begin{table}[htbp]
    \centering
    \caption[Study~11: Aggregate Shapley by source]{ Study~11: Shapley attributions per source for the ten-candidate cohort in \ref{tab:shapley-verification}.
    The shares sum positive $\phi_i$ only, excluding the negative GitHub penalty from the pie's demoninator to avoid the potential for zero division,
    for future evaluation runs.}
    \label{tab:shapley-source-dominance}
    \vspace{1.5ex}
    \small
    \begin{tabular}{l|c|c}
        \hline
        \textbf{Source} & \textbf{Sum of positive $\phi_i$} & \textbf{Share of pool (\%)} \\ \hline
        CV (claimed)       & 2.429 & 59.6 \\
        GitHub (demonstrated) & 1.564 & 38.4 \\
        LinkedIn (claimed)    & 0.085 & 2.1 \\ \hline
    \end{tabular}
\end{table}

Across the cohort, demonstrated GitHub evidence supplies a large share of positive attribution despite sparse 
LinkedIn personas. This is shown in Table~\ref{tab:shapley-source-dominance} and Figure~\ref{fig:shapley-source-dominance} 
(Appendix~\ref{appendix:shapley-charts}).

\subsection{Study 14: HCI Trial}
\label{subsec:hci-trial-protocol}

Study 11 mathematically confirms that the explainable AI (XAI) engine is consistent, and explains the practical utility
of these innovations. However, we are yet to demonstrate the human-facing elements of the XAI engine; this is addressed by Study 14. 
Since the purpose of our XAI engine is to be integrated into a front-end application, we must measure how
MERIT's "Glass-Box" design influences recruiter behaviour, builds cognitive trust, and mitigates any potential biases.

The objective of this study is to measure both the decision and confidence calibration of recruiters, along with the usability
of the explainability layer. We cover three central hypotheses:

\begin{itemize}
    \item \textbf{Hypothesis 1 (Decision Calibration):} Access to the Glass-Box XAI audit trail will increase the rate at 
    which recruiters reverse incorrect screening decisions for adversarial candidates (such as
          keyword-stuffers and identity squatters) and "hidden gems" (like high-potential candidates with low tenure).
    \item \textbf{Hypothesis 2 (Confidence Calibration):} Recruiters will exhibit a higher alignment between their
          screening confidence and the empirical evidence, resulting in increased confidence for verified candidates and
          decreased confidence for unverified profiles.
    \item \textbf{Hypothesis 3 (Usability and Trust):} The interactive explainability layer will achieve a high System
          Usability Scale (SUS) score and receive high qualitative ratings for transparency and defensibility compared to a
          traditional black-box score.
\end{itemize}

\subsubsection{Two-Stage Experimental Protocol}
We start with a two-stage user study involving professional technical recruiters and hiring managers as participants.
This trial follows a two-stage process for each candidate in the pool:

\subsubsection{Stage 1: Control Condition (Black-Box Score)}
Recruiters are given a candidate CV with a single aggregate match alongside per-metric scores, but with no explanations.
Recruiters must make a decision on whether to shortlist or reject the candidate, and rate their confidence on a Likert scale
(1 = Not Confident at all, 5 = Extremely Confident). They must also provide a brief qualitative justification for their decision.

\subsubsection{Stage 2: Treatment Condition (Glass-Box Scorecard)}
Recruiters are now shown the same candidate, but with access to the glass-box scorecards, including the Shapley layer, audit trail,
and other explainability features. Recruiters will then be asked to state whether they wish to reverse or maintain their decision.
They are also asked to rate their confidence, along with an explanation of how the visualisations influenced their assessment, if at all.

\subsubsection{Key Quantitative Metrics and Empirical Results}
This pilot user trial includes 5 participants, consisting of 2 professional recruiters ($R_1, R_2$) with extensive experience in
software development talent acquisition, and 3 technical peers ($P_1, P_2, P_3$), representing MEng Computing
students at Imperial College London; three of these peers were acquainted with the author. 
These participants are familiar with the technical recruitment process and have undergone multiple technical projects and internships. 
Given $N=5$, we treat this as a formative pilot and report descriptive statistics, inferential tests are 
exploratory only (Section~\ref{sec:threats-to-validity}).

To prevent participant fatigue, we present each evaluator with only 4 synthetic candidates to evaluate. Unlike the personas
mentioned in previous sections, these candidates were designed in a real-world scenario. We used real candidate information
as the basis for synthesising these candidates:
\begin{itemize}
    \item \textbf{Moinecha Ramos-Horta} (Verifiable Elite Developer)
    \item \textbf{Yusuf Nilsson} (The "Hidden Gem" - 1 year of experience on CV, elite GitHub)
    \item \textbf{Robert Todd} (Adversarial Keyword Stuffer)
    \item \textbf{Aaron Mohammed} (Identity Squatter - claims deep Java/Go systems expertise on CV, squatted GitHub)
\end{itemize}

We present the quantitative results of this trial in Table~\ref{tab:hci-trial-results}. Decision-reversal, confidence-shift, and SUS distributions are in Appendix~\ref{appendix:hci-trial-charts} (Figures~\ref{fig:hci-decision-reversals}--\ref{fig:hci-sus-distribution}). Raw session data are in \texttt{evaluation/14-hci\_trial/}.


\begin{table}[htbp]
    \centering
    \caption[Study~14: HCI pilot results]{ Quantitative results of Study~14, with $N = 5 Evaluators$. Decisions are split into shortlisted and rejected decisions ($S$ and $R$ respectively).
    Decision Reversal Rate (DRR) refers to a Stage 1 decision that was later reversed in Stage 2 after being presented with the Glass-Box scorecard.
    Confidence follows the 5-point Likert scale (1 = Not Confident, 5 = Extremely Confident)}
    \label{tab:hci-trial-results}
    \vspace{1.5ex}
    \small
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{l|c|c|c|c|c|c}
            \hline
            \textbf{Candidate}   & \splithead{Stage 1 Decision}{(Control)} & \splithead{Stage 2 Decision}{(Treatment)} & \splithead{Reversal Rate}{(DRR)} & \splithead{Avg. $C_{\text{Stage 1}}$}{(1--5)} & \splithead{Avg. $C_{\text{Stage 2}}$}{(1--5)} & \splithead{Confidence Delta}{($\Delta C$)} \\ \hline
            Moinecha Ramos-Horta & $5S$ / $0R$                             & $5S$ / $0R$                               & 0.0\%                            & 3.80                                          & 4.60                                          & +0.80                                      \\
            Yusuf Nilsson        & $0S$ / $5R$                             & $4S$ / $1R$                               & 80.0\%                           & 4.20                                          & 4.80                                          & +0.60                                      \\
            Robert Todd          & $4S$ / $1R$                             & $1S$ / $4R$                               & 75.0\%                           & 3.60                                          & 4.40                                          & +0.80                                      \\
            Aaron Mohammed       & $3S$ / $2R$                             & $0S$ / $5R$                               & 100.0\%                          & 3.20                                          & 4.80                                          & +1.60                                      \\ \hline
        \end{tabular}%
    }
\end{table}

This pilot provides directional support for all three hypotheses under realistic hiring constraints, 
none are confirmed at inferential strength given $N=5$. We detail the results below:

\subsection*{Quantitative Results}

\subsubsection{Decision Calibration (Hypothesis 1)}
During the first stage, we can see the issue of the "resume black hole", all 5 candidates rejected Yusuf Nilsson
due to his low tenure (1 year of experience on CV). However, upon transitioning to stage 2, we see that 4/5 evaluators
(80\% DRR) reversed their decision, recognising his valuable github contributions. Recruiter $R_2$ chose to
maintain the rejection, noting that the JD strictly required 3+ years of experience regardless
of technical strength. This illustrates how human decision-makers calibrate system outputs with non-quantifiable
operational constraints.

Another key finding is that the explainability layer successfully aided in unmasking adversarial manipulation. More specifically,
we turn to the candidate \textbf{Robert Todd}, who managed to fool 4 out of 5 evaluators in stage 1 due to his keyword-dense
resume and aggregate score. However, upon transitioning to stage 2, we see that 3 out of 4 of those evaluators (75\% DRR)
reversed their decision, recognising his keyword stuffing.

The most significant instance of this is with the candidate \textbf{Aaron Mohammed}, who managed to fool 3 out of 5 evaluators
in stage 1 due to his gamified CV. However, upon transitioning to stage 2, rather than seeing supported evidence, recruiters were
presented with a significant red flag. Aaron Mohammed provided a GitHub profile that was not their own in order to pass off
someone else's work as his own. This is a serious ethical breach and a violation of professional integrity. Due to this, all
3 evaluators reversed their decision, rejecting Aaron Mohammed. It is crucial to note that recruiters did not reject
Mohammed based on the fact that he was not a good developer (in fact, according to his CV he was an expert in Java and Go).
Instead, they rejected him based on the fact that he showed a lack of integrity and dishonesty. This illustrates how the explainability layer
can be used to identify and reject participants who are not a good fit for the company culture, even if they are technically skilled.


\subsubsection{Confidence Calibration (Hypothesis 2)}
Once evaluators were presented with the Glass-Box MERIT interface, we observed a positive confidence shift ($\Delta C$)
across all profiles. This was particularly notable in the case of Moinecha Ramos-Horta, where her average confidence
rose from $3.80$ (Stage 1) to $4.60$ (Stage 2), an increase of $0.80$. In a similar fashion, confidence for Aaron Mohammed
also increased, rising from $3.20$ (Stage 1) to $4.80$ (Stage 2), an increase of $1.60$, which was
due to the veto penalty that stemmed from his dishonesty, giving the evaluators the confidence to reject him with high conviction.

The average confidence across all 20 profile-level observations (5 participants $\times$ 4 candidates) increased from $3.75$ 
(Stage 1) to $4.45$ (Stage 2), a mean increase of $0.70$ on a 5-point scale. A paired $t$-test on these rows yields 
$t(19) = -5.60$, $p < 0.001$, but the rows are not independent (fixed Stage~1$\rightarrow$Stage~2 order, repeated participants, 
acquaintance bias among peers). We therefore report this as descriptive evidence for Hypothesis~2, not confirmatory inference.

\subsubsection{Usability (Hypothesis 3)}
To evaluate the usability of MERIT's glass-box architecture, we followed the standard System Usability Scale (SUS) survey \cite{brooke1996sus}.
MERIT achieved an average SUS score of \textbf{81.50} out of 100, with individual scores ranging from 77.5 to 85.0.
According to the Bangor et al. acceptability scale, this corresponds to an ``Excellent'' rating (Grade A) at face value.
With $N=5$, it indicates interface feasibility only, not product-market validation (Section~\ref{subsec:construct-validity}).

\subsubsection{Qualitative Trust \& Feedback Analysis}
The qualitative feedback gathered from the evaluators highlighted how MERIT's explainability layer resolves the standard
friction of automated screening.

One of the technical recruiters ($R_1$) noted:
\begin{quote}
    \textit{"Traditionally, ATS scores are black boxes. We either trust them blindly or ignore them completely. Seeing exactly
        how much GitHub contributed versus self-reported CV claims completely changed how I assessed Yusuf Nilsson. It gave me
        the confidence to shortlist a candidate I would have normally filtered out within seconds."}
\end{quote}

Professional Recruiter $R_2$ added a comment reflecting realistic operational friction:
\begin{quote}
    \textit{"While the Glass-Box scorecard shows he has elite technical skills verified by public code, our client's visa
        requirements strictly mandate a minimum of 3 years of professional tenure. I cannot shortlist him."}
\end{quote}

Similarly, MEng Student Peer $P_3$ noted:
\begin{quote}
    \textit{"The keyword stuffing warning is a serious red flag, but the resume still claims senior Go and Kubernetes
        experience. I will keep him shortlisted for a brief technical phone screen to probe his actual knowledge."}
\end{quote}

Technical Recruiter $R_1$ added regarding the identity mismatch veto:
\begin{quote}
    \textit{"The second I saw the red 'Identity Mismatch Veto' flag on Aaron's profile, it was an immediate reject. In our
        job, honesty is everything. If the software shows you've linked a GitHub account that is active under a completely
        different name, there is no point looking any further. It saved me a lot of wasted phone screening time."}
\end{quote}

Collectively, these results provide formative support for MERIT's explainability pillar. Study~11 confirms theoretical consistency. 
This pilot suggests the layer may aid practical, human-in-the-loop review, but a larger recruiter panel is needed before stronger claims.
