\section{Fairness, Security \& Computational Rigour}
\label{sec:fairness-computational-rigour}

The previous sections covered the accuracy of MERIT, its fusion logic, explainability and also it's alignment with human experts.
We now move on to the last stage of our evaluation by testing the fairness, security against adversarial inputs and computational 
scalability of MERIT. Refer to Appendix~\ref{appendix:fairness-scalability-charts} for 
and \ref{appendix:bias-anonymisation-charts} for the charts associated with this section.

\subsection{Study 15: Bias Mitigation \& Anonymisation Efficacy}
\label{subsec:bias-mitigation-efficacy}

The previous sections established that MERIT is able to accurately rank candidates based on their CV and public-repository evidence.
It is also able to explain its decisions, and align with human experts. We now move on to the fairness of MERIT, by testing 
it's ability to mitigate bias and anonymise candidate data. In this study (\texttt{15-bias\_anonymisation\_audit/}), we will be testing 
the efficacy of Blind Mode, a UI feature which redacts PII and other sensitive information from the candidate's data. Blind Mode
also standardises the layout of the candidate's data, including the CV to ensure a fair comparison between candidates, devoid
of any visual bias.

The aim of this study is to test whether Blind Mode on MERIT's per-candidate Intelligence Report popup reduces the reliance
on visible proxy markers, compared to the default presentation that exposes the candidate's CV and PII. Evaluators did
not see the four-candidate ranking overlay (closing the popup would have revealed it). They did still see each candidate's match score and 
Glass-Box breakdown in the popup. We use a between-subjects design so each evaluator only sees one UI condition and is not affected by 
an earlier visible review of the same people. Two hypotheses guide the pilot:

\begin{itemize}
    \item \textbf{H1 (Proxy-gap compression):} The shortlist-rate gap between prestige-high and prestige-mid/low proxy groups 
    will be smaller in the Blind arm than in the Visible arm.
    \item \textbf{H2 (Title and layout debiasing):} Candidates penalised by intern titles, noisy parsing, or role-skewed visible 
    layout (e.g. Vincent) will have higher shortlist rates in the Blind arm, with MERIT scores held constant.
\end{itemize}


\subsubsection{Experimental protocol}
In our experiment, we have four recruiters ($r1$, $r2$, $r3$, $r4$) and four peer evaluators ($p1$, $p2$, $p3$, $p4$). 
The participants in this study were in a different panel from Study~14, so they have no prior participation, or interaction with
MERIT. Evaluators were assigned in balanced blocks to ensure a fair comparison (2 recruiters and 2 peers per block).

\begin{itemize}
    \item \textbf{Block 1: Visible arm} ($n = 4$, B1 $\in$ \{\textbf{R1, R2, P1, P2}\}): Intelligence Report popup with Blind Mode 
    \textbf{off} (names, employers, institutions, titles, standard layout).
    \item \textbf{Block 2: Blind arm} ($n = 4$, B2 $\in$ \{\textbf{R3, R4, P3, P4}\}): same popup with Blind Mode \textbf{on by default} 
    (candidate identifiers, redacted institutions with tier retained, neutralised skill layout. 
    Section~\ref{sec:bias-mitigation-safeguards}).
\end{itemize}

For fairness, all participants reviewed the same four profiles in a fixed order against the Intern Backend Engineer JD from 
Study~07. We only present recruiters with the per-candidate Intelligence Report popup, ensuring that the ranking overlay 
is not visible to them. Within the popup, however, each candidate's MERIT match score and Glass-Box skill breakdown remained 
visible in both arms. This is a constraint of the product, rather than an intentional protocol choice. After viewing this popup, 
evaluators were prompted to record their decision (Shortlist/Reject), confidence (1--5), and a brief justification.
No multi-source evidence was presented to the participants, and the scores were identical across arms, in other words, the engine
was never changed. No evaluator saw both Visible and Blind UI on the same candidates.

\paragraph{Design rationale.}
This protocol aims to test the blind redaction on singly-profile reviews, disallowing them from seeing that 
candidate's rank in the batch. Post-hoc analysis does compare the human shortlist rates against the MERIT ranks, 
but those ranks were not displayed during the session. We would not allow the evaluators to view the 
cohort rankings to avoid potential bias or memory effects. The between-subjects design removes the effect of 
visible-then-blind memory on the same evaluator. We have that $n = 4$ per arm due to the limited number of participants,
this increases the variance of the results, but we are still able to draw relatively reliable conclusions. Results are directional only.

\subsubsection{Cohort design}
Synthetic candidates were selected from Studies~07 and~08, with no overlap with the Study~14 set 
(Moinecha Ramos-Horta, Yusuf Nilsson, Robert Todd, Aaron Mohammed):
\begin{itemize}
    \item \textbf{Aisling Hunt} (Study~07) --- junior data scientist, UC Berkeley, feminine name and gendered 
    extracurricular signals on the visible CV.
    \item \textbf{Vincent Germain} (Study~08) --- intern backend engineer, human rank~1 in Study~08 but intern title 
    and corrupted PDF header on the visible view.
    \item \textbf{Riko Al-Tayer} (Study~08) --- senior engineer, Cambridge/Imperial/Stripe pedigree (prestige-halo 
    risk, overqualified for an intern JD).
    \item \textbf{Jude Washington} (Study~07) --- junior iOS developer, Berkeley education but unknown employers 
    (TechFlow, Momentum), and a mobile-first visible layout.
\end{itemize}

As expected, MERIT Full scores are unchanged by Blind Mode (UI-only redaction): Riko (60.9\%) $>$ Jude (39.9\%) $>$ Aisling (36.7\%) $>$ Vincent (31.1\%).

\subsubsection{Quantitative results}
Table~\ref{tab:bias-audit-summary} reports how often each candidate was shortlisted in each arm ($n = 4$ evaluators per arm). 
Figure~\ref{fig:bias-shortlist-rates} and Figure~\ref{fig:bias-blind-uplift} (Appendix~\ref{appendix:bias-anonymisation-charts}) 
plot the same data. Table~\ref{tab:bias-disparity-metrics} summarises the prestige proxy gap used for H1.

\begin{table}[htbp]
    \centering
    \caption{Blind Mode audit summary (Study~15, between-subjects: $n = 4$ per UI arm, 4 candidates). Rates are the 
    fraction of evaluators in that arm who shortlisted the profile.}
    \label{tab:bias-audit-summary}
    \vspace{1.5ex}
    \small
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{l|c|c|c|c}
            \hline
            \textbf{Candidate} & \textbf{MERIT (\%)} & \textbf{Visible arm (\%)} & \textbf{Blind arm (\%)} & \textbf{$\Delta$ (pp)} \\ \hline
            Riko Al-Tayer      & 60.9 & 100.0 & 75.0 & $-$25.0 \\
            Aisling Hunt       & 36.7 & 25.0  & 75.0 & +50.0 \\
            Vincent Germain    & 31.1 & 0.0   & 75.0 & +75.0 \\
            Jude Washington    & 39.9 & 25.0  & 50.0 & +25.0 \\ \hline
        \end{tabular}%
    }
\end{table}

\begin{table}[htbp]
    \centering
    \caption{Aggregate disparity metrics (Study~15, between-subjects). Prestige-high: Riko and Aisling; prestige-mid/low: Vincent and Jude.}
    \label{tab:bias-disparity-metrics}
    \vspace{1.5ex}
    \small
    \begin{tabular}{l|c|c}
        \hline
        \textbf{Metric} & \textbf{Visible arm} & \textbf{Blind arm} \\ \hline
        Evaluators (recruiters / peers) & 2 / 2 & 2 / 2 \\
        Mean shortlist rate --- prestige-high group (\%) & 62.5 & 75.0 \\
        Mean shortlist rate --- prestige-mid/low group (\%) & 12.5 & 62.5 \\
        Prestige gap (high minus mid/low, pp) & 50.0 & 12.5 \\ \hline
    \end{tabular}
\end{table}

\paragraph{Visible arm ($n = 4$).}
Riko was shortlisted by all four evaluators (100\%). Vincent received no shortlists (0\%): recruiters cited his intern title, 
short tenure, and a corrupted PDF header on the visible CV. Aisling and Jude each received one shortlist (25\%). In this arm, 
decisions clearly followed visible prestige cues (university names, employer logos) and job titles more than backend-intern fit.

\paragraph{Blind arm ($n = 4$).}
\textbf{H1} is supported in direction: the prestige gap fell from 50~pp to 12.5~pp (Table~\ref{tab:bias-disparity-metrics}), 
mainly because Vincent (0\% $\rightarrow$ 75\%) and Aisling (25\% $\rightarrow$ 75\%) were shortlisted far more often. 
\textbf{H2} is strongest for Vincent and Aisling---the profiles most penalised by visible layout and titles. Riko dropped to 
75\% (3/4): one recruiter rejected him as overqualified for an intern role even though he had the highest MERIT score. Jude 
rose to 50\% (2/4) once employers were hidden and skill bullets were easier to read. Blind Mode did not make everyone agree 
with MERIT's ordering; it changed \emph{who} got a fair first look when names and schools were removed.

\subsubsection{Qualitative themes}
Facilitator notes (\texttt{responses.txt}) match the table. In the visible arm, evaluators mentioned firm logos, school names, and 
``intern'' labels. In the blind arm, they referred to skill lists and whether the person fit the intern JD. Match scores and 
breakdowns were always shown in the popup, so recruiters could still lean on MERIT's number. We cannot claim they reviewed 
in isolation from the engine. The four-candidate ranking overlay was never shown.

\subsubsection{Limitations}
Due to the small sample size, we cannot draw any statistical conclusions from the data. We can only observe the trends and 
direction of the results. We also cannot rule out the possibility that the evaluators were influenced by the visible scores and 
breakdowns, which would bias the results. Shortlist rates can only move in steps of 25\% because each arm only had four voters,
and MERIT scores were always visible in the popup and could act as an underlying anchor.

\subsubsection{Study 15 takeaway}
Blind Mode is useful as a first-pass safeguard, it reduces obvious proxy bias in this pilot and helps profiles like 
Vincent get considered when titles and formatting had put them at a disadvantage. It does not replace Study~14's Glass-Box 
checks, and it is not proof of demographic fairness.

\subsection{Study 06: Adversarial Stress Testing}
\label{subsec:study-06-adversarial}

This study provides a small-scale evaluation of MERIT's reaction to a base candide that has been duplicated with 
different pertubations to simulate different real-world scenarios and adversarial attacks. All candidates in this test are 
a copy of the same base candidate, \textbf{Honest Candidate}, and the pertubated candidate's name is often derived from
the scenario or attack they represent. \textbf{Honest Candidate} is a junior software engineer with a strong CV and a fair 
match percentage of 49.8\%. We have the following scenarios:
\begin{itemize}
    \item \textbf{Ghost}: A candidate that has keyword stuffed their public-repository to appear more skilled than they are.
    \item \textbf{Fraud}: A candidate that has inflated their skills to appear more skilled than they are.
    \item \textbf{Stale}: A candidate that has updated their CV to appear more skilled than they are, but their public-repository evidence is still outdated.
    \item \textbf{Gamer}: A candidate that has manipulated their CV and public-repository to appear more actively engaged in particular metrics.
    \item \textbf{Squatter}: A candidate that has squat the name of another candidate to appear more skilled than they are. 
    In this case, the candidate was caught by MERIT's Integrity Layer.
    \item \textbf{Smart Squatter}: A candidate that has injected the name of another candidate to appear more skilled than they are.
    In this case, the candidate was able to bypass the Integrity Layer by finding an appropriate name that would pass the identity audit,
    bypassing the fuzzy name-match. Note that this is a rare scenario in the real-world.
    \item \textbf{Shadow}: A candidate that has no public-repository evidence to support their claims (as opposed to Ghost,
    who mainly has contradictory evidence, but still has some evidence).
    \item \textbf{Inflater}: A candidate that has inflated their repository metrics to appear more skilled than they are.
\end{itemize}

Each scenario is scored by both the CV-only and Full MERIT engines. 
Results are exported to \texttt{output/multi\_source\_adversarial\_matrix.csv}. 
Table~\ref{tab:adversarial-matrix} summarises the scores. 
Figures~\ref{fig:adversarial-cv-vs-full} and~\ref{fig:adversarial-robustness} 
(Appendix~\ref{appendix:fairness-scalability-charts}) compare CV-only versus 
Full MERIT per scenario and report deviation from the honest baseline.

\begin{table}[htbp]
    \centering
    \caption{Adversarial robustness matrix (Study~06). All personas share the same \textbf{Honest Candidate} baseline (CV-only 29.6\%, Full 49.8\%). 
    $\Delta_{\mathrm{honest\_cv}}$ is each scenario's CV-only score minus the honest CV-only score (29.6\%). 
    $\Delta_{\mathrm{fusion}}$ is Full MERIT minus CV-only on the same row. 
    $\Delta_{\mathrm{honest\_fusion}}$ is each scenario's Full MERIT score minus the honest Full score (49.8\%). 
    Figure~\ref{fig:adversarial-robustness} visualises $\Delta_{\mathrm{honest\_fusion}}$ with attack-level annotations.}
    \label{tab:adversarial-matrix}
    \vspace{1.5ex}
    \footnotesize
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{l|l|c|c|c|c|c}
            \hline
            \textbf{Scenario} & \textbf{Persona} & \textbf{CV (\%)} & \textbf{$\Delta_{\mathrm{honest\_cv}}$} & \textbf{Full (\%)} & \textbf{$\Delta_{\mathrm{fusion}}$} & \textbf{$\Delta_{\mathrm{honest\_fusion}}$} \\ \hline
            Honest            & Honest Candidate   & 29.6 & 0.0   & 49.8 & +20.2 & 0.0   \\
            Ghost             & The Ghost          & 26.7 & $-$2.9  & 34.8 & +8.1  & $-$15.0 \\
            Fraud             & The Fraud          & 44.3 & +14.7 & 47.4 & +3.1  & $-$2.4  \\
            Stale             & Stale Candidate    & 35.0 & +5.4  & 37.9 & +2.9  & $-$11.9 \\
            Gamer             & The Time Traveler  & 35.0 & +5.4  & 58.1 & +23.1 & +8.3  \\
            Squatter          & Vince Vault        & 29.6 & 0.0   & 29.8 & +0.2  & $-$20.0 \\
            Smart Squatter    & Rayan Akhtar Smith & 29.6 & 0.0   & 49.8 & +20.2 & 0.0   \\
            Shadow            & The Shadow         & 29.6 & 0.0   & 30.4 & +0.8  & $-$19.4 \\
            Inflater          & The Inflater       & 29.6 & 0.0   & 46.1 & +16.5 & $-$3.7  \\ \hline
        \end{tabular}%
    }
\end{table}





\textbf{Successful Behaviours.}
The \textbf{Squatter} scenario is the key highlight of the study, showing that the Integrity Engine can successfully detect and penalise most
forms of identity squatting. As shown, CV-only is unchanged, and fusion only adds an insignificant uplift of $+0.2$. As shown through ($\Delta_{\mathrm{honest\_fusion}}$),
There is a $20$ point difference between two identical candidates, highlighting that a candidate's score can be significantly impacted by MERIT if they attempt to game 
the system. Since both candidates have the same CV-only score, the difference is solely due to the Integrity Engine.

Another successful response to a  scenario is the \textbf{Stale} scenario, in which a candidate attempts to game the system by updating their CV to appear more 
skilled than they are, but their public-repository evidence is still outdated. As shown, CV-only is unchanged, and fusion only adds an insignificant uplift of $+2.9$. 
Relative to the honest candidate, the score is reduced by $-11.9$ points, highlighting that outdated skills are not fully rewarded once GitHub is included.
In this case, it is important to note that they are not penalised, but rather that the effects of Bayesian Fusion update the Beta Distribution to reflect the lower
signal strength from GitHub.

\textbf{Ghost} is also another response that we believe to be successful, in which a candidate attempts to game the system by keyword stuffing their CV 
while also providing a public-repository that contradicts their claims. As shown, CV-only partialy penalised due to the keyword-stuffing. What is more important
lies in the $\Delta_{\mathrm{honest\_fusion}}$ score, which is $-15.0$. This is due to the Bayesian Fusion updating the Beta Distribution to reflect the lower
signal strength from GitHub as they are unable to provide evidence of their skills.

\textbf{Partial mitigations.}
An example of a partially successful instance is the \textbf{Fraud} scenario, in which a candidate successfully inflates their CV-only score through false claims,
while providing repository ecidence that contradicts this. Unlike \textbf{Ghost}, the candidate overhauls their CV to appear more skilled than they are, rather than
simply keyword stuffing their CV. As shown, CV-only is increased by $+14.7$ points. When it comes to the Bayesian Fusion, the score is increased by $+3.1$ points.
This is due to the Bayesian Fusion updating the Beta Distribution to reflect the low signal strength from the candidate's GitHub profile.

\textbf{Documented failure modes.}
The bigget failure with the Integrity Engine, namely, the \textbf{Smart Squatter} scenario. In this case, the candidate was able to bypass the integrity engine by
finding a GitHub portfolio that matches the candidate's name, is able to pass the fuzzy name-match, and is able to score a high score when evaluated against that
specific JD. This is a rare scenario in the real-world, but it is important to note that it is possible.

Another impractical failure scenario is the \textbf{Gamer} scenario, in which a candidate attempts to game the system through the manipulation of public-repository metrics.
In this scenario, the candidate successfully achieves a high GitHub score ($\Delta_{\mathrm{fusion}} = +23.1$). In this case, the candidate was able to successfully trick 
the system into thinking that they were recently active in particular metrics required for the role. This highlights a critical flaw in the way we calcualte the year to 
project temporal skill decay.


\subsubsection{Study 06 takeaway}
The integrity engine is able to detect and penalise most forms of realistic adversarial attacks, such as keyword stuffing, identity squatting, and repository gaming.
Note that the majority of failure modes are edge-cases that take significant effort to achieve, and are not expected to be a major concern in the real-world.
For example, lets consider the \textbf{Smart Squatter} scenario. For a single role, lets assume the malicious candidate wishes to squat a GitHub portfolio. First,
they would need to find portfolios that would likely score high on MERIT, which is a relatively trivial task, but is still considered a significant effort.
From here, the actor would need to find a name that is close enough to bypass the fuzzy name-match, this is likely not a guaranteed success due to the rarity of
full names that match. This effort would technically work, but in the real-world, where a candidate applies to a vast array of jobs with different metrics, 
this would be impractical. In our example, we consider GitHub portfolios, but this identity squatting can be applied to other data sources, such as LinkedIn as well.

\subsection{Studies 02 \& 03: Computational Scalability \& Performance Benchmarks}
\label{subsec:studies-02-03-scalability}
The next two studies (Study~02 and Study~03) focus on the performance of MERIT, in terms of both runtime and spacetime complexities for a batch of 
$N \in \{10, 50, 100, 200, 500\}$ candidates. We will benchmark these against Traditional ATS, Modern AI ATS, and MERIT. To truly measure the differences
in runtime and spacetime, we separate MERIT into three configurations: CV-Only, Full multi-source, and Full with Shapley explainability. Study~02
initialises each engine once, and records the wall-clock latency for an entire batch. Study~03 on the other hand, measures memory in a fresh subprocess per
engine (static load) and per (engine, $N$) batch size. This ensures that RSS is never affected by other engines resident in the same process, at the cost 
of a longer benchmark run. Static figures are RSS growth from engine initialisation through one warmup score.

Benchmarks were executed in a single run on Windows~11 (AMD Ryzen~7 7800X3D, AMD Radeon RX~7900~XT, Python~3.11.9). Scoring is CPU-bound and does not use 
GPU acceleration. Results measure the scoring engine only. It does not account for PDF ingestion or network I/O.


\subsubsection{Time complexity and latency (Study 02)}
All engines run in $O(N)$ time complexity, with each additional candidate adding roughly the same amount of work, however, with different constant factors.
Table~\ref{tab:runtime-results} and Figure~\ref{fig:runtime-complexity} summarise these results. For graphical
representations of the data, please refer to Appendix~\ref{appendix:fairness-scalability-charts}.

\begin{table}[htbp]
    \centering
    \caption{Batch ranking latency (Study~02). Wall-clock seconds to score an entire cohort.}
    \label{tab:runtime-results}
    \vspace{1.5ex}
    \small
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{r|c|c|c|c|c}
            \hline
            \textbf{$N$} & \textbf{Traditional ATS (s)} & \textbf{Modern AI ATS (s)} & \textbf{MERIT CV-Only (s)} & \textbf{MERIT Full (s)} & \textbf{MERIT Explainable (s)} \\ \hline
            10  & 0.0035  & 4.50   & 0.095 & 0.107 & 0.528 \\
            50  & 0.0127  & 21.86  & 0.471 & 0.486 & 2.678 \\
            100 & 0.0231  & 43.70  & 0.916 & 0.983 & 5.521 \\
            200 & 0.0489  & 88.05  & 1.797 & 1.976 & 10.92 \\
            500 & 0.1059  & 219.29 & 4.579 & 4.808 & 26.79 \\ \hline
        \end{tabular}%
    }
\end{table}

At the largest batch size ($N = 500$), we can observe that Modern AI ATS requires $\approx$219.3 seconds to score the batch,
meanwhile MERIT Full with Shapley explainability requires $\approx$26.8 seconds to score the batch. This means that MERIT, even 
with the explainability layer, runs $\approx 8.18$ times faster than Modern AI ATS at this scale. It is important to note the overhead
of the explainability layer too, which is $\approx 5.57$ times slower than MERIT Full (with no explainability). This result is to be expected
due to the overhead of the shapley value computations. Currently with $3$ sources, we need to compute $2^3 = 8$ coalitions, which is a relatively small number.
However, as we add more sources, it is important to note that this approximate slowdown will increase. More specifically, for $N$ sources, we need to compute $2^N$ coalitions, 
which is a factorial increase in the number of computations. This is a critical limitation of the current implementation, and is a topic for future work.
Another key observation is the difference in runtime between MERIT CV-Only and MERIT Full. With a batch of 500, we only see a $0.229$ second difference, making
the effects of multi-source fusion relatively minimal.


\subsubsection{Space complexity and memory footprint (Study 03)}
Study~03 separates \textbf{static engine footprint} (model and interpreter load at initialisation) from 
\textbf{operational batch overhead} (incremental RSS growth while scoring a batch, after garbage collection). 
Table~\ref{tab:engine-load} and Table~\ref{tab:spacetime-batch} summarise \texttt{output/engine\_load\_costs.csv} and 
\texttt{output/spacetime\_results.csv}; Figure~\ref{fig:spacetime-composite} (Appendix~\ref{appendix:fairness-scalability-charts}) visualises both.

\begin{table}[htbp]
    \centering
    \caption{Static engine memory footprint (Study~03). Per-engine RSS 
    delta from initialisation through one warmup score, each measured in a 
    fresh subprocess (MB).}
    \label{tab:engine-load}
    \vspace{1.5ex}
    \small
    \begin{tabular}{l|c}
        \hline
        \textbf{Engine} & \textbf{Static Load (MB)} \\ \hline
        Traditional ATS   & 0.10 \\
        Modern AI ATS     & 396.2 \\
        MERIT CV-Only     & 366.3 \\
        MERIT Full        & 367.1 \\
        MERIT Explainable & 367.2 \\ \hline
    \end{tabular}
\end{table}

\begin{table}[htbp]
    \centering
    \caption{Incremental batch memory overhead (Study~03). Per-(engine, $N$) RSS growth during batch 
    scoring in a fresh subprocess, measured after \texttt{gc.collect()} with the engine already 
    loaded (MB). Same engine ordering as Table~\ref{tab:runtime-results}.}
    \label{tab:spacetime-batch}
    \vspace{1.5ex}
    \small
    \resizebox{\textwidth}{!}{%
        \begin{tabular}{r|c|c|c|c|c}
            \hline
            \textbf{$N$} & \textbf{Traditional ATS (MB)} & \textbf{Modern AI ATS (MB)} & \textbf{MERIT CV-Only (MB)} & \textbf{MERIT Full (MB)} & \textbf{MERIT Explainable (MB)} \\ \hline
            10  & 0.01 & 114.3 & 54.4 & 54.5 & 53.5 \\
            50  & 0.02 & 115.8 & 56.6 & 56.8 & 56.5 \\
            100 & 0.03 & 116.6 & 58.7 & 58.8 & 59.3 \\
            200 & 0.04 & 120.1 & 64.1 & 64.1 & 64.5 \\
            500 & 0.10 & 120.6 & 73.5 & 74.9 & 76.2 \\ \hline
        \end{tabular}%
    }
\end{table}

Under isolated subprocess measurement, static load is negligible for Traditional ATS (\textbf{0.1~MB}) but dominated 
by transformer-backed stacks for Modern AI ATS (\textbf{396.2~MB}) and MERIT (\textbf{366--367~MB}), reflecting SBERT 
and backend registry materialisation on the warmup score---still modest in absolute terms for a desktop deployment. 
At batch scale, Traditional ATS remains near zero; Modern AI ATS holds a near-flat incremental footprint 
($\approx$\textbf{114--121~MB}) from embedding workspaces. MERIT configurations scale roughly linearly with cohort size 
(\textbf{54~MB} at $N = 10$ to \textbf{76~MB} at $N = 500$) because the harness retains per-candidate audit payloads; 
Explainable is highest at large $N$. Multi-source fusion adds little RSS over CV-Only, however, the explainability layer is the 
main memory premium within MERIT.


\subsubsection{Studies 02 \& 03 takeaway}
Batch latency can be observed to scale linearly with $N$ for all engines, where $N$ refers to the number of candidates in a batch.
MERIT Full scores the largest measured batch size ($N = 500$) in under $5$ seconds, whereas Modern AI ATS requires $\approx$219 seconds to score the same batch.
Introducing Shapley values introduces a $\approx 5.6 \times$ runtime increase with three sources ($2^3 = 8$ coalitions), this is due to the overhead of the
shapley value computations. The exponential cost is in the number of evidence sources, not the cohort size. The biggest risk to merit is if many data sources
are introduced. In a real-world context, this isn't a significant concern, but it is important to note that this could become a potential risk if the recruitment
process decides to introduce more data sources to the screening process. The memory overhead of MERIT is shown to be quite negligable after accounting for this, 
as through the difference in RSS growth between MERIT Explainable between $N = 10$ and $N = 500$ is only $22.7MB$. This is a relatively small amount of memory, 
and is likely to be negligible in the context of a production system.

\medskip
\noindent
The purpose of this section is not to re-validate the mechanisms of MERIT, but to instead understand whether the system is mature enough to be deployed in 
a profuction environment, under proxy biases, manipulation and scale. Study~15 helped to show that MERIT is able to mitigate bias and anonymise candidate data,
Study~06 later showed that we can block most forms of naive attempts to game both previously existing ATS systems, as well as MERIT itself with documented
residual gaming risks. Studies~02 and ~03 both highlighted that MERIT is both spacetime and runtime efficient, with explainability being the main memory premium,
and Shapley values being the main runtime premium. MERIT is best positioned as an auditable screening assistant that is fast enough for batch triage,
but also transparent enough for recruiter override. It is also anonymisable enough to reduce some first-pass proxy noise, serving as a fair and responsible system.
MERIT should not be considered as a fully autonomous or fairness-certified hiring authority, but rather as a tool to aid the recruitment process. 