\section{Human-Machine Alignment}
\label{sec:human-machine-alignment}

The previous rank-based accuracy metrics measured the performance of MERIT against specific adversarial candidates
with simplistic synthetic data. This was performed in order to demonstrate significant changes in rank orderings based 
on the candidates' persona. In this section, we will instead have recruiters rank a set of realistic synthetic candidates
in order to evaluate the overall performance of MERIT's scoring system against a fixed expert consensus.

This Section will use the Spearman rank-correlation test between the rankings produced by MERIT and the expert consensus.
We will also compare the performance of MERIT against two CV screening baselines (Traditional ATS and Modern AI ATS) in order 
to act as a baseline for comparison. Throughout this chapter, we will present numeric tables for each study. To see 
a graphical representation of the results, please refer to Appendix~\ref{appendix:spearman-charts}.

\subsection{Methodology}
\label{subsec:spearman-method}

We will run four different studies, each with a different set of ten candidates as well as a job description to match that
cohort against. The ten candidates are synthetic individuals created to be realistic and representative of the target cohort.
We create synthetic JSON data for their CV, GitHub and LinkedIn profiles to be used as input for the scoring engines. 
The candidate CV is used as a template for synthetic data generation, and is also kept as a source of truth for the expert consensus.

Throughout this section, \textbf{Traditional ATS} and \textbf{Modern AI ATS} label in-repository CV screening baselines (keyword matching and SBERT cosine similarity, respectively) that model scoring stages commonly embedded in rule-based and AI-assisted Applicant Tracking Systems (ATS).
They are not commercial ATS products. MERIT is a multimodal screening system evaluated alongside these baselines.

To emulate the real-world hiring process, we will have between one and three recruiters decide on a rank ordering of the candidates.
For this study, to ensure realistic results, we do not anonymise the candidates, and instead use their real names.
The consensus is then encoded as an ordered list of CV filenames in \texttt{human\_rankings.txt}, where rank 1 is the best match
for the job, while rank 10 is the worst.

For each engine, we feed in the Job Description and candidate CVs to produce a rank ordering of candidates, based on 
descending order of the match score outputted by that engine. In our tests, $h_i$ denotes the human rank, while $e_i$
denotes the engine rank for a given candidate $i$. We can then calculate the Spearman rank-correlation coefficient $\rho$
with the following formula:
\begin{equation}
    \rho = 1 - \frac{6 \sum_i (h_i - e_i)^2}{n(n^2 - 1)} .
\end{equation}

We report the correlation coefficient and two-sided $p$-value from the standard test implemented in the evaluation harness 
(\texttt{scipy.stats.spearmanr}), with $n = 10$ for all four studies. With only ten ranks, $\rho$ is sensitive to a single swap; $p$-values are reported for completeness but should be read as exploratory. The four engines are compared on identical inputs,
like with the previous studies on candidate personas. Once again, the engines are as follows: \textbf{Traditional ATS}, 
\textbf{Modern AI ATS}, \textbf{MERIT (CV Only)}, and \textbf{MERIT (Full)} (CV plus LinkedIn and GitHub enrichment).

\subsection{Study 07: High Discrimination}
\label{subsec:study-07-spearman-high-discrimination}

The first study will be a high discrimination scenario to find the best fit for a full-stack backend engineer. Our recruiters 
can clearly see a significant distinction between candidates due to there being a diverse pool of candidates with different 
seniorities and also career paths. Human experts can clearly separate clear fits from weak matches. 
Figure~\ref{fig:spearman-studies-combined} (panels a--d, one per study) summarises $\rho$ for each engine, 
while the underlying coefficients are recorded in Table~\ref{tab:spearman-study-07}. 

\begin{table}[htbp]
    \centering
    \caption[Study~07: Spearman $\rho$ (High Discrimination Cohort)]{Study~07: Spearman $\rho$ for the high-discrimination cohort.}
    \label{tab:spearman-study-07}
    \small
    \setlength{\tabcolsep}{5pt}
    \begin{tabular}{l|c|c}
        \hline
        \textbf{Engine} & \textbf{Spearman $\rho$} & \textbf{$p$-value} \\ \hline
        Traditional ATS & $0.5515$ & $0.0984$ \\
        Modern AI ATS   & $0.2848$ & $0.4250$ \\
        MERIT (CV Only) & $0.5152$ & $0.1276$ \\
        MERIT (Full)    & $0.6970$ & $0.0251$ \\ \hline
    \end{tabular}
\end{table}

In this study, MERIT tracks the expert ordering better than the baselines, however, when only considering CV information,
it performs worse than the AI baseline, but still better than the traditional ATS. The exported \texttt{detailed\_rankings.csv} 
confirms that the top human pick (Moinecha Ramos-Horta) is ranked first by MERIT (Full), while weak matches are pushed down the list.


\subsection{Study 08: Seniority-Bias}
\label{subsec:study-08-spearman-seniority}

The next study will be a seniority-bias stress test to find the best fit for an intern backend engineer. Here, our recruiters
are more likely to rank interns higher than seniors due to the junior nature of the role. Usually more experienced candidates would
get ranked lower, due to a perceived assumption that they are overqualified for the role. This nuance goes unaddressed by all four
engines since this is a rare edge-case in real-world scenarios. It is not often that a senior candidate would apply for an intern role.

\begin{table}[htbp]
    \centering
    \caption[Study~08: Spearman $\rho$ (seniority bias)]{Study~08: Spearman $\rho$ for the seniority-bias cohort.}
    \label{tab:spearman-study-08}
    \small
    \setlength{\tabcolsep}{5pt}
    \begin{tabular}{l|c|c}
        \hline
        \textbf{Engine} & \textbf{Spearman $\rho$} & \textbf{$p$-value} \\ \hline
        Traditional ATS & $-0.6606$ & $0.0376$ \\
        Modern AI ATS   & $-0.8061$ & $0.0049$ \\
        MERIT (CV Only) & $-0.5152$ & $0.1276$ \\
        MERIT (Full)    & $-0.5152$ & $0.1276$ \\ \hline
    \end{tabular}
\end{table}

For this cohort, the synthetic GitHub and LinkedIn profiles are intentionally designed to align with the given candidate's CV.
None of the profiles introduce contradictory evidence. MERIT (Full) raises the absolute scores of candidates, but still preserves the same rank orderings as MERIT (CV Only)
due to this alignment of evidence. This does not lead to the conclusion that multi-source fusion is ineffective, but rather that multi-source fusion is discriminatory
upon the presence of contradictory evidence.

Inspecting the rankings (found in \texttt{detailed\_rankings.csv}), 
the human-preferred intern profile (Vincent Germain, human rank~1) is ranked quite low across all scoring engines.
On the other hand, principal/staff CVs score higher, highlighting the gap between MERIT's current weighting and this
scenario's ``potential over title'' consensus. The outcome of this is quite valuable as an honest stress test, as opposed 
to a cherry-picked success case. Here, we see that all scoring engines are unable to understand the nuances of candidate
recruitment when it comes to overqualified candidates.


\subsection{Study 09: Peer Competition}
\label{subsec:study-09-spearman-peer}
After conducting studies on cases with significant differentiators such as seniority and career path, we now move on to 
a more closely matched pool of candidates. The aim of this study is to find the best fit for a mid-level backend engineer.
It is important to note that in this study, recruiters have solely based their rankings on CV information, and not on 
any other sources of information (LinkedIn, GitHub) to ensure a fair comparison.

As shown in Table~\ref{tab:spearman-study-09}, the traditional ATS achieves the 
strongest alignment with the expert consensus, while MERIT (Full) achieves the second highest alignment. It is worth noting the
vast difference in alignment between MERIT (CV Only) and MERIT (Full), which is a clear indication that the additional sources of information
are doing substantial work to keep full MERIT in the same broad ordering as experts. The presence of external evidence can help to 
act as a tie-breaker when the CV-only scores are too close to call.

\begin{table}[htbp]
    \centering
    \caption[Study~09: Spearman $\rho$ (peer competition)]{Study~09: Spearman $\rho$ for the peer-competition cohort.}
    \label{tab:spearman-study-09}
    \small
    \setlength{\tabcolsep}{5pt}
    \begin{tabular}{l|c|c}
        \hline
        \textbf{Engine} & \textbf{Spearman $\rho$} & \textbf{$p$-value} \\ \hline
        Traditional ATS & $0.7939$ & $0.0061$ \\
        Modern AI ATS   & $0.2364$ & $0.5109$ \\
        MERIT (CV Only) & $0.0909$ & $0.8028$ \\
        MERIT (Full)    & $0.6242$ & $0.0537$ \\ \hline
    \end{tabular}
\end{table}

We can observe tight score clustering in the rankings, which explains the weak 
CV-only $\rho$. The additional sources of information are able to act as a tie-breaker when the CV-only scores are too close to call.
While this case doesn't highlight MERIT (Full) as the best performing engine, it is still able to achieve a strong alignment with
the expert consensus and put forward a strong case for the use of additional sources of information to help with tie-breaking.

\subsection{Study 10: Signal Dissonance}
\label{subsec:study-10-spearman-dissonance}

The next study is a cherry-picked failure case to highlight the Achilles' heel of MERIT (Full). Note that the setup for this study
is almost identical to Study~07, with the key difference being the introduction of conflicting signals between the CV and 
external evidence. This is done to induce high dissonance between the public-repository evidence and CV narrative, corrupting
the ability of MERIT (Full) to correctly rank the candidates.

Table~\ref{tab:spearman-study-10} shows that MERIT (Full) achieves a
near-total loss of rank alignment with the expert consensus, with a $\rho$ of 0.0061 and a $p$-value of 0.987. By using 
the exact same setup as Study~07, but with the introduction of conflicting external signals, we are able to pinpoint the 
failure of MERIT to one specific vulnerability: \emph{the introduction of contradictory multi-source evidence}. This sends a 
clear message to the reader that MERIT (Full) can only perform as well as the quality of the evidence it is given.


\begin{table}[htbp]
    \centering
    \caption[Study~10: Spearman $\rho$ (signal dissonance)]{Study~10: Spearman $\rho$ for the signal-dissonance cohort.}
    \label{tab:spearman-study-10}
    \small
    \setlength{\tabcolsep}{5pt}
    \begin{tabular}{l|c|c}
        \hline
        \textbf{Engine} & \textbf{Spearman $\rho$} & \textbf{$p$-value} \\ \hline
        Traditional ATS & $0.5515$ & $0.0984$ \\
        Modern AI ATS   & $0.2848$ & $0.4250$ \\
        MERIT (CV Only) & $0.5152$ & $0.1276$ \\
        MERIT (Full)    & $0.0061$ & $0.9867$ \\ \hline
    \end{tabular}
\end{table}

In the detailed rankings,
we can see that the worst-pick candidate (Jude Washington, human rank~10) is promoted to the top spot by MERIT (Full), while the 
top-pick candidate (Moinecha Ramos-Horta, human rank~1) is dropped to the second spot. This was an intentional design choice to
make the failure case more obvious and easier to understand. We clearly show that if the data sources are contradictory, 
the multi-source pillar of MERIT collapses and is unable to make confident and accurate decisions.

\subsection{Conclusion}
\label{subsec:spearman-synthesis}

Across the four studies, no single engine dominates every scenario, which is expected, given the nature of the studies and 
the different scenarios they are designed to test. MERIT (Full) delivers the best statistically supported agreement
in high-discrimination scenarios, but can collapse under adversarial dissonance. It's key to note that MERIT's multi-source
pillar is dependent on the quality of evidence provided, and relies upon candidates to provide accurate and consistent evidence.

While candidates are free to provide contradictory evidence, in a real-world scenario, this is unlikely to happen, as it is unreasonable
to provide contradictory evidence on their CV and public-repository evidence. If candidates have inconsistent narratives between their 
CV and public-repository evidence, this would be a clear sign of a candidate that is not being honest and transparent about their work.