\section{Threats to Validity}
\label{sec:threats-to-validity}

The evaluation in this chapter is designed to be both comprehensive and rigorous, however, no empirical study
is without its limitations. In this section, we cover any threats to validity that may affect the validity of
our results. Following the established taxonomy of Wohlin et al.~\cite{wohlin_threats}, we follow the same categorisations.
We cover construct validity, internal validity, external validity and conclusion validity. Across these fifteen evaluation 
studies, we attempt to provide evidence of the systemic behaviour of MERIT's architecture, whilst also maintaining 
transparency regarding the limitations of these studies.


\subsection{Construct Validity}
\label{subsec:construct-validity}

Construct validity refers to the extent to which the metrics and instruments used in the evaluation studies genuinely
measure the research questions that are being investigated.

\textbf{Self-Defined Ground Truth and Researcher-Authored Personas.}
Most studies rely on ground truth being authored by myself. For example the parser labels and synthetic JSON profiles
and sabotaged GitHub payloads. This presents the risk of confirmation bias, where evaluation data may inadvertently encode
the behaviours that the ending was designed to exhibit. We aim to mitigate this through intentionally including deliberate
failure cases to avoid the unintentional cherry-picking of successful cases. The key example of this is Study~10, where we 
deliberately sabotage GitHub payloads to deliberately contradict the CV narrative to show the flaws of MERIT (Full). Note that
we do not claim independent external validation of ground truth datasets due to the constraints of the project.

\textbf{Proxy Metrics for Recruitment Quality.}
The evaluation measures parser F1-scores, rank correlations, shortlist rates, SUS and batch latency. Note how these metrics
are not downstream hiring outcomes such as interview pass rates and offer acceptance. A high Spearman $\rho$ or a narrow
Blind Mode prestige gap does not indicate that MERIT (Full) is able to improve hiring decisions. We would need to perform
field trials with outcome labels to actually assess the effectiveness of MERIT (Full) in improving hiring decisions.

\textbf{Prestige Proxies Versus Demographic Fairness.}
For Study~15, we test the fairness of Blind Mode through layout standardisation, redaction of names, employers and academic institutions,
while still retaining the tier labels. It is critical to note that the redaction of some fields, such as names, can help to ensure
fairness, however, other fields, such as employers and academic institutions, are not protected characteristics. Therefore, 
the redaction of these fields does not guarantee fairness. Blind Mode must not be read as a fairness certificate.

\textbf{Automation Bias and Visible Engine Scores (Study~15).}
As mentioned in Section~\ref{subsec:bias-mitigation-efficacy}, we did
not mask the engine scores and glass-box breakdowns in the Blind Mode arm. This means that evaluators may 
have anchored on the engine scores and glass-box breakdowns rather than re-deriving judgments from redacted text alone.
For example, evaluators may have been more likely to shortlist candidates with high engine scores, which could lead to a bias
in the results.

\textbf{Stress-Testing CV--GitHub Contrast.}
MERIT is designed to treat LinkedIn as a supplemental source alongside the CV. LinkedIn arrives as structured API data, so it is not subject to
the same PDF parsing errors as CVs (Studies~04--05). When a CV is incomplete or parsed incorrectly, LinkedIn can help recover claimed skills and
employment history without repeating the extraction pipeline. That said, LinkedIn is still self-reported evidence, much like the resume. GitHub is
different, as it reflects demonstrated work in public repositories.

In Studies~06 and~07--10, we did not use ``full'' LinkedIn personas (rich history, endorsements, and skills mirroring the CV). The cohorts rely on
clean synthetic CV JSON, so there is no parser failure for LinkedIn to fix, and a dense profile would mostly repeat claims already in the resume.
We therefore kept LinkedIn sparse on purpose, to stress-test the contrast we cared about: CV narrative against GitHub evidence, without a second
copy of the same self-report diluting the signal. We still included minimal LinkedIn fields so the three-source pipeline and Study~11 Shapley
decomposition remain valid (for example, null-player checks), but this is not the same as recruiters reviewing a complete LinkedIn profile in
production. We should be careful when generalising fusion and Shapley results from these studies to live hiring where LinkedIn is fully populated
(Section~\ref{subsec:study-11-shapley-verification}).


\textbf{Usability and Trust Instruments (Study~14).}
The HCI study is discussed in Section~\ref{subsec:hci-trial-protocol},
SUS here, indicates the prototype's feasibility, rather than product-market validation.

\subsection{Internal Validity}
\label{subsec:internal-validity}
Internal validity concerns whether the observed results are genuinely caused by the independent variables under study, 
rather than by confounding factors.

\textbf{Simulated Baseline Systems.}
The ``Modern AI ATS'' and ``Traditional ATS'' baselines serve as comparators to MERIT (Full) and MERIT (CV-Only). They are
not proprietary commercial products, nor open-source alternatives. They are both in-repository baselines under identical local inputs.
Reported speed-ups (e.g.\ $\approx 46\times$ at $N=500$) compare algorithm families upon the author's workstation. As such, it does
not account for vendor SLAs, API latency, or managed-cloud deployments. Traditional ATS is keyword-counting only and may understate
modern vendor semantic features.

\textbf{Single-Run and Single-Machine Benchmarks.}
Studies~02--03 record one wall-clock and memory sweep per configuration on a single consistent development workstation. The specifications
of this workstation are discussed in Section~\ref{subsec:studies-02-03-scalability}. Note
that there are no confidence intervals, cross-hardware replication and repeated trials. There is a likelihood of shifts in terms of 
absolute seconds or MB figures on rerun. We deem these shifts to be relatively insignificant due to the general pattern of runtime and
memory overhead being relatively consistent across the different configurations.

\textbf{Fixed HCI Protocol Order (Study~14).}
In the HCI study, Stage~1 always precedes Stage~2, and as such, the reversals and confidence gains may partly reflect order effects rather
than explainability alone.

\textbf{Adversarial Scenario Construction.}
Study~06 clones a baseline `Honest Candidate' across eight different personas, cleanly isolating perturbation effects. However, it does
not account attack sophistication across a real applicant pool. The residual failures, such as the `Smart Squatter' and `Gamer' scenarios,
are documented as high-effort edge cases, but their base rate in live hiring is unknown.

\textbf{Cherry-Picked Failure and Success Cases.}
Studies~01 and ~10 both include intentionally engineered failure and success narratives in order to aid explaination. However, such
narratives may shadow the general pattern of the engine's performance, and as such, they may not be representative of the general population.
Please refer to Section~\ref{subsec:study-10-spearman-dissonance} and 
Section~\ref{subsec:study-01a-algorithmic-ablation} for more details.

\subsection{External Validity}
\label{subsec:external-validity}

External validity covers the extent to which our findings may be generalised beyond the conditions of these studies.

\textbf{Small Cohort Sizes and Discrete Outcomes.}
Due to the small cohort sized of $n=10$ candidates per JD, the Spearman rank correlation coefficient $\rho$ is quite volatile,
and $p$-values are also sensitive to one rank swap. Because of this, we cannot make any claims about the generalisability of the results.
In a similar vein, Study~15 employs $n=4$ evaluators per arm, and four candidates, so shortlist rates move in 25\% steps, and
cannot support most claims. Study~14's $N=5$ and Study 13's $N=4$ cannot represent global recruiter populations.

\textbf{Open-Source and Public-Evidence Bias.}
MERIT's multi-source pillar benefits from the use of public evidence, such as public GitHub respositories and LinkedIn profiles.
Candiates such as Carl Corp introduce a `Proprietary Data Gap' (Section~\ref{subsec:study-01b-data-source-ablation}) 
due to the lack of public evidence. Candidates in this situation may be unfairly penalised relative to juniors with active GitHub
histories, introducing a structural bias toward open-source visibility that may correlate with access to candidate data.

\textbf{Evaluation--Production Gap.}
Benchmarking measures do not account for the latencies of PDF upload, network I/O, rate-limited API calls to GitHub/Linkedin
and database persistence. Production deployments on shared servers or containers are likely to differ from the 
isolated subprocess RSS on the author's desktop workstation.


\subsection{Conclusion Validity}
\label{subsec:conclusion-validity}

Conclusion validity assesses the extent to which the conclusions drawn from the data are justified. We also consider
whether the methods used are appropriate.


\textbf{Spearman's Rank Correlation as the Primary Metric.}

The usage of Spearman $\rho$ measures the monotonic agreement with expert ordering. It does not measure the calibration of 
match percentages, nor the magniture of the rank swaps. Engines with a similar $\rho$ may produce significantly different score 
spreads. Therefore, we report $\rho$ alongside the detailed rankings available in the GitHub repository this project is stored in,
to allow for inspection, rather than treating the spearman correlation coefficient alone as sufficient.

\textbf{Interpretation of Negative Correlations (Study~08).}
The negative $\rho$ under the intern JD stress test (Section~\ref{subsec:study-08-spearman-seniority}) 
indicates misalignment rather than sampling noise. We do not claim seniority-aware hiring without explicit policy rules.

\textbf{Incomplete Empirical Validation of Shapley Axioms.}
Study~11 verifies the axions of efficiency, null-player, additivity and determinism on ten synthetic profiles. The axiom of symmetry,
however, is omitted since the `players' of this cooperative game are intentionally heterogeneous. Negative GitHub attributions,
such as Vince Vault, are mathematically valid in the game we study, however, they could be cognitively surprising.


\textbf{Multiple Comparisons Across Studies.}
We have conducted fifteen studies on overlapping personas and engines. This presents the risk of researcher degrees of freedom,
where the researcher may selectively emphasise supportive studies and document failures. We present failure modes explicitly to 
counter narrative cherry-picking, but the chapter should not be read as a meta-analysis with pooled effect sizes.

\subsection{Mitigations \& Bounded Claims}
\label{subsec:validity-mitigations}

Despite the limitations we note above, this chapter still provides convergent evidence about how MERIT behaves under controlled
data. We cannot treat this as a proof of production readiness, however, we can frame three bounded claims that follow directly from
the studies.

\textbf{The use of evidence.}
In the emulated JSON cohorts (Studies~01, 06, 07--10), MERIT's multi-source fusion and integrity checks are shown to move 
rankings in directions that match the design's intent. An example of this is Jordan Smith, a candidate who is usually penalised 
by CV-only systems, however, is uplifted when GitHub is included. Another example is the lowering of scores when presented with
evidence that contradicts the candidate's claims.


\textbf{Glass-Box and Blind Mode affordances.}
In the HCI studies (Studies~14--15), both the Glass-Box overlay, as well as Blind Mode can influence the recruiter's behaviour,
as well as their stated confidence. We treat these as formative findings only, given the convenience samples and visible match scores
in the popup for Blind Mode.


\textbf{Prototype operability.}
For the benchmark studies (Studies~02--03), MERIT is shown to remain within sub-minute targets for $N \le 500$ with the full
capabilities of MERIT (all sources and full explainability). The overall pattern of runtime and memory overhead is still
enough to support a feasibility argument for batch screening.

We do not claim that MERIT is fairness-certified, nor do we with to present it as a fully autonomous hiring authority.
Future work would be needed to address these limitations, such as counterbalanced HCI trials, larger independent recruiter panels,
repeated benchmarks with variance reporting, end-to-end PDF ingestion tests, and field studies linking MERIT outputs to interview
and hire outcomes. MERIT is best described as an \textbf{auditable screening assistant}. It's purpose is to help recruiters
inspect evidence and override the engine if needed, rather than replacing human judgement outright.

