\section{Future Work}
\label{sec:future-work-roadmap}

The suggested future work follows from the evaluation gaps listed in earlier sections, we order these suggestions by their
potential for both hiring harm and deployability. They are not ordered by implementation convenience, and instead
focus on the most impactful changes that would help MERIT to be a more useful tool for recruiters.



\subsection{Evidence Ecosystem \& Missing-Data Fairness}
\label{subsec:future-evidence-ecosystem}

\subsubsection{Motivation}
As shown in Study~01B, we have that multi-source fusion rescues strong candidates that have weak CVs, but we
also see the same logic punishing engineers whose best work is private. MERIT swaps one bias for another, in this case
we see the introduction of open-source visibility bias.

\subsubsection{Additional Sources}
Chapter~\ref{sec:data-sources} already argues for platforms such as LeetCode, HackerRank,
and Stack Overflow and complementary proficiency signals. A natural extension is a plug-in ingestion interface
that normalises each API into MERIT's evidence schema. Coding platforms are particularly valuable for candidates 
with sparse GitHub histories, but strong contest or assessment records. We could also use the platform's data on 
attempted questions to infer a candidate's proficiency in a given language or problem domain.

\subsubsection{Missing Data Policy}
A good missing data policy could be to distinguish three states per source: \emph{present-positive}, 
\emph{present-negative/conflicting}, and \emph{unknown}. An empty data source (such as an empty GitHub profile)
should never tank a candidate's score, but should instead be down-weighted to avoid significant penalties.
Private-repository contributions could be supported through signed attestations or verified employer 
metadata rather than public commit counts only. Suggestions such as the following could prevent MERIT
from penalising candidates such as Carl Corp.



\subsection{Robust Ingestion \& Role Intelligence}
\label{subsec:future-ingestion}

\subsubsection{Motivation}
Pillar~1, as shown in prior sections, is significantly limited by layout-driven PDF failures, as well as JD noise.
Extensibility built on mis-extracted skills may propagate error into every downstream rank.

\subsubsection{Layout-robust Parsing}
Benchmark-driven resume models (such as ResumeBench \cite{emnlp2025resumebench}) should replace or augment 
our current coordinate-based extraction \textit{pdfminer} \cite{pdfminer}, with an evaluation arm that mirrors
Study~04 on live applicant PDFs, as opposed to template-only corpora. The acceptance criterion should be explicit,
multi-column experience recall within a fixed margin of single-column performance before fusion is trusted on CV text alone.

\subsubsection{Role Intelligence}
Study~13's cross-role IDF interference suggests that we should maintain separate corpora per role family (backend, frontend, data, etc.)
so that common languages (such as Python) can appear as valuable in roles where it is not the status quo. Skill families should be 
distinguishing, for example, React front-end from Node back-end under a shared ``JavaScript'' label. This would allow for more 
accurate metric suggestions that reflect sub-domain requirements rather than frequency in a pooled corpus.




\subsection{Fusion \& Conflict Handling}
\label{subsec:future-fusion}

\subsubsection{Motivation}
Study~10 demonstrates that MERIT has the capability to fail completely when presented with a batch full of candidates 
that have drastically conflicting evidence. While this is a rare occurance, it is a clear indication that MERIT's multi-source pillar
is not robust enough to handle conflicting evidence. Fusion without safeguards is worse than CV-only screening.

\subsubsection{Evaluation Cohorts}
Fusion and Shapley studies use sparse LinkedIn personas to isolate the CV--Github contrast. A replication with fullo
populated LinkedIn profiles should be conducted before generalising source-attribution results to production,
where LinkedIn often duplicated, but also corrects, CV claims.

\subsubsection{Product Behaviour}

The dashboard should surface explicit conflictions, such as when a candidate's CV asserts seniority in Java,
but their GitHub profile shows zero Java commits in the past year. There should be the application of confidence
gating. When inter-source variance exceeds a threshold, the engine should prompt for human review, rather than 
produce an expected score from the beta distribution we use. Integrity checks such as the ones in Study~06 should be evolved
with new gaming patterns, such as candidate's attempting to have multiple accounts on different platforms, tailored to 
different industries and roles.

\subsection{Explainability at Scale}
\label{subsec:future-explainability-scale}

\subsubsection{Motivation}
The current implementation of data-source attribution is accurate, but costly to compute at scale.
While the cost does not scale beyond a linear scale for candidates, we cannot say the same for the way MERIT 
scales with the number of data sources. The $O(2^n)$ complexity is a critical limitation and a topic for future work.
Alongside this, the hard-coded permutations in our code base will need accounting for if the above suggestion is implemented 
(LeetCode, HackerRank, Stack Overflow, etc.). It is infeasible and inappropriate to hard-code permutations for all possible 
combinations of sources in a real-world context, however, for the purposes of this project, we believe it to suffice due
to the small number of sources we currently support.


\subsubsection{Approximation and Caching}
For $n \ge 5$, we should migrate to Monte Carlo, or a similar approximation technique to compute the Shapley values.
Recruiters rarely need full, exact attributions on every applicant, perhaps only on shortlists. A dual-mode architecture would
not be a bad consideration, with rank-only mode for bulk batches, and audit properly for the top-$k$ candidates, with
dynamic coalition generation replacing static source lists.

\subsubsection{Blind and Contestable UI}
Future HCI work should test blind mode once engine scores are masked to ensure fairness. We must also counterbalance the 
stage order too, as well as using independant panels. Explainability should default to highlighting conflict and uncertainty,
rather then only positive skill bars. This way, glass-box mode encourages calibrated scepticism, rather than give recruiters
an excuse to fall upon automation bias.



\subsection{Validation Beyond the Prototype}
\label{subsec:future-validation}

\subsubsection{Motivation}
While the parser $F_1$, Spearman~$\rho$, and SUS show the prototype behaves as designed, this does not claim improved hire
quality. Deployment claims require evidence backed by outcome-linked evidence.

\subsubsection{Field Studies}
The step of largest value that we could take is a longitudinal study that links MERIT rankings to interview
pass rates, offer acceptance, and (where ethically and legally permissible), early on-the-job performance.
A small pilot with one hiring partner, pre-registered hypothesis and blinded human review would 
allow MERIT to move away from architecture validation and instead towards recruitment science.

\subsubsection{Rigorous Usability and Benchmarks}
Study~14's HCI trial is a good starting point for usability and confidence calibration. We should also consider
using larger independent recruiter panels, as well as repeated Studies~02--03 with variance bands across hardware.
We should also compare against commercial ATS APIs under identical inputs where licensing allows. All such work
must operate within GDPR constraints on personal hiring data (Section~\ref{sec:data-privacy-gdpr}).

Until these steps exist, MERIT should remain as an auditable assistant that helps recruiters
inspect evidence and challenge the engine where necessary. Once again, we reiterate that MERIT is not a 
fairness-certified or hire-optimising product.