\chapter{Related Work \& Technical Foundations}

Following the identification of systemic biases and cognitive limitations in the traditional technical recruitment process, this chapter explores the technical frameworks designed to automate and standardise candidate evaluation.
The analysis starts off with a review of the current state-of-the-art in Applicant Tracking Systems (ATS), categorising them based on their underlying logic.
By examining the limitations of these systems, specifically their lack of transparency, as well as reliance on a single source of data, this chapter establishes a technical requirement for multi-source explainable architectures.
This technical foundation justifies the multimodal, evidence-driven design of MERIT~\cite{merit_repo} as an essential advancement in the field of automated technical recruitment.

\input{background/subsections/automated_screening.tex}

Researchers have explored a variety of methods for aiding technical recruitment, particularly in the area of automated screening.
MERIT will focus on three representative strands of work that have helped to influence its design.

\section{Multi-Agent LLM Architectures for Explainable Screening}
There have been a variety of approaches when it comes to the automation of CV screening. One recent example is Lo et al.\ \cite{lo2025aihiring}, 
who describe in an arXiv preprint a multi-agent LLM architecture that splits the screening process into four distinct 
stages: resume-extraction, evaluation, summarisation and score-formatting. The authors take a modular approach to the screening process,
ensuring a clear separation of concerns between the different stages of the screening process.

To improve contextual relevance, the system integrates Retrieval-Augmented Generation (RAG) within the evaluator agent,
 enabling the use of external knowledgeable sources (such as certifications and university rankings) rather than relying solely on the content of the CV itself.

In their evaluation, the authors compare the AI-generated candidate scores with ratings by human HR professionals on a dataset of anonymised resumes.
The results demonstrate that the multi-agent RAG-LLM system achieved strong agreement with human judgements and 
outperformed simpler baselines, indicating its potential to replicate, and perhaps improve on, the traditional screening workflows at scale.

However, a significant limitation remains: while the architecture is more flexible than rule-based ATS, the 
evaluation still relies on a predetermined set of scoring dimensions (skills, experience, education, background and summary), 
each of which has fixed weights.
The fixed dimensionality hinders the system's ability to accurately reflect role-specific priorities, unique 
project requirements, unconventional career paths, or evolving technical criteria that may fall outside of standard categories.
As a result, the framework struggles to fully capture the diversities and nuances inherent in real-world hiring.

To conclude, while the paper suggests that LLM-based modular screening is feasible and 
can approximate human performance, it is still challenged with balancing benchmarked consistency with adaptive, 
context-sensitive evaluation.

\section{Technical Role Inference from Open-Source Contributions}

Research also demonstrated the merits of going beyond the CV to understand a potential candidate's technical background.
In this case, we consider Montandon et al.\ \cite{montandon2021_roles}. The authors propose a supervised machine learning
approach to infer six technical roles through the usage of GitHub profiles and project metadata (e.g.\ programming languages,
repository descriptions and short bios). The authors constructed a manually labelled dataset of roughly 2{,}284 developers to
train and evaluate several classification models. Their best models report an aggregate precision of 0.88 and an AUC of 0.89,
with programming language features being the most discriminatory factor in this classification process.
Note that recall is weaker for some roles, most particularly for DevOps and mobile.

The authors of this paper believe that this approach should be considered a supplementary hiring signal rather than
the sole criterion for hiring, and understandably so. However, the study remains quite limited for recruitment
screening in a few ways that are relevant to this project. The first limitation concerns regex matching of
profile text, which is prone to self-report bias. Public GitHub activity may also not reflect a candidate's skill set,
especially for candidates who contribute primarily to private repositories. Alongside this, the role classifications
lack refinement, and are far too coarse and fixed, lacking the extensibility required for a system that needs to
process candidates from a variety of different roles and specialisations. The misclassifications that the authors
report also highlight how reliance on only repository-language signals can misinterpret a developer's specialisation.
This work does not cross-check other sources of data, such as the CV, for external evidence, and nor does it
provide explainable or justifiable reasoning for why a candidate was predicted to hold a particular role.

Regardless of these limitations, the paper motivates MERIT to treat GitHub as a supplementary signal beyond the CV.
MERIT will not adopt this classifier directly, but will integrate GitHub to corroborate or
challenge a candidate's claims on their CV.

\section{Semantic Embedding for Latent Skill Extraction}
Research aimed at improving job-candidate matching has also moved beyond keyword-based methods.
Gugnani and Misra \cite{gugnani2020} propose a system that identifies both the explicit and implicit skills 
from job descriptions by combining traditional NLP techniques with large-scale document embeddings.
The authors of this research leveraged approximately one million job postings to train a vector embedding model that
attempts to capture the semantic relationships between the job description, and the skills required for that role.
It attempts to uncover hidden, context-dependent skills that the recruiter may not have originally considered, allowing
for the screening process to have more differentiating capabilities with regards to candidate evaluation.


A core limitation of this research, however, ties in with the rigid nature of the scoring system.
The candidate-job matching relies on a fixed set of weightings, consistently prioritising explicit 
skills and integrating similarity, frequency, and implicit-skill penalties according to predefined 
coefficients. The rigid scoring system limits its adaptability across roles that vary in how skills are expressed, be it implicit, or explicit.
Also, the quality and reliability of inferred skills are tightly coupled to the representativeness 
of the training corpus, meaning that gaps and biases in the job posting can influence recommendations.

Despite these constraints, the study effectively shows how document-level semantic models can 
complement manual and keyword-based extraction approaches, enhancing the detection of latent skills. 
It also highlights the challenges in designing scoring mechanisms that remain flexible across diverse occupations.
As automated job recommendation systems increasingly rely on inferred competencies, this shows 
the importance of developing adaptive, domain-sensitive models that are capable of capturing the 
variations in how industries emphasise or omit particular skills.


\input{background/subsections/bayesian_fusion.tex}
\input{background/subsections/shapley_values.tex}
\input{background/subsections/semantic_embeddings.tex}

\input{background/subsections/relevance_to_merit.tex}
