\chapter{Declarations}
\label{appendix:declarations}

\section{Use of Generative AI}

Throughout the development of MERIT, we have used generative AI tools and models (ChatGPT, Cursor, Google Antigravity IDE, Claude Sonnet 3.5, and Gemini Pro)
for routine implementation support: frontend styling, boilerplate code, test scaffolding, and syntax-level
debugging. We have also leveraged them for minor proofreading and \LaTeX{} formatting of the report. All factual
claims, numerical results, and technical arguments were written and verified by the author.

All personas and scenarios were defined by the author, which were then used to generate the synthetic evaluation dataset.
LLM-assisted tools aided in the generation of synthetic candidate JSON and job-description corpora. We reviewed,
edited and validated all files before use. Where we have used real candidate data for the generation of synthetic profiles,
the author obtained consent from the individuals involved, and ensured that their data was not present in the synthetic dataset.

The core algorithm design, such as Bayesian Fusion and scoring logic, are the author's original work. This extends to the 
system architecture, evaluation design, and also the interpretation of results. In instances where an AI tool suggested a 
technical approach (specifically, game-theoretic source attribution), we have independently reviewed the 
literature, and only adopted the approach upon verification.

All submitted work reflects the author's understanding and responsibility for the project's design, implementation, and evaluation.

\section{Ethical Considerations}

This project involved processing sensitive data from candidate CVs, LinkedIn profiles, and GitHub repositories. 
Ethical issues such as privacy, consent, and bias were carefully considered. 
Most candidate data used during testing was anonymised or synthetic to prevent disclosure of personal information.
For live LinkedIn ingestion tests, at most 30 profiles were retrieved via Apify, each belonging to a personal contact 
who gave informed consent for FYP use only (see Chapter~6). This approach is not intended for deployment.
Where live data were stored, they were held in Supabase with access restricted via environment variables, no sensitive 
data were published in the repository.

\section{Sustainability}

The project does not pose any significant environmental impacts.
MERIT is lightweight in terms of carbon emissions and resource consumption, 
it does not require any significant infrastructure or resources to run.


\section{Availability of Data and Materials}
Refer to the following link for the availability of data and materials: \url{https://github.com/RayanAkhtar/MERIT}.
This repository contains source code, evaluation studies, and supporting scripts.
Note that the full synthetic evaluation dataset is not included in this repository, we only include the relevant
datasets for the evaluation studies. This dataset can be found in the \textit{/test\_data} directory in a given evaluation study.
Full access has been provided to allow replication, evaluation and the verification of results.
With the sole exception of the author, there is no sensitive or personally identifiable information that has been shared or stored in this 
repository.
