\section{Ethical AI \& Hiring}
\label{sec:ethical-ai-hiring}

As we have seen in previous sections, Automated Screening Tools (ATS) have the potential to repeat old biases, hide candidate 
outcomes, and also encourage candidates to game the system \cite{bogen2018,raghavan2020,fabris2025fairness}. Most ATS products
are labelled as ``black boxes'', meaning that there is little transparency into the decision-making process. MERIT, on the 
other hand shows why a score was produced. The open question however, is whether recruiters leverage this transparency to ensure 
fairness, or only use it to defend a decision that they had already intended on making.

In Study~14, we observed that the visibility of the Glass-Box breakdown impacted behaviour. Recruiters overturned shortlists
when evidence of keyword stuffing and identity squatting were spotted. Note that in the case of identity squatting, recruiters would
reject candidates not on the basis of their skills, but rather on evidence of dishonesty. While this is a positive outcome, we must also
note that we cannot be certain that recruiters would have overturned the shortlist without the Glass-Box breakdown. As stated in Study~15,
match scores are intentionally provided in Blind Mode, however, this could introduce some bias. Similarly, the inclusion of the tier labels
of the candidate's academic institution could still introduce some bias.

The Integrity Engine aided in blocking most simple cheats, as evident in Study~06, though clever gaming of activity metrics and rare name-match
identity squatting are still possible, and valid concerns should be raised when using MERIT in the hiring process. Study~15 was also 
informative on the effectiveness of Blind Mode in reducing bias. We were able to observe that unfair biases were mitigated through 
the design choices made in MERIT. This suggests a fairer first look, but not a proof of fairness across gender or ethnicity. 
Note that Bertrand and Mullainathan's field experiments documented name-based discrimination in hiring callbacks \cite{bertrand2004}, 
while Study~15 did not test those protected characteristics directly.

MERIT's evidence-first approach leads to the system being more biased towards candidates who have public work samples, such as GitHub.
This is evident in Study~01b, where Jordan Smith, a candidate who is usually penalised by CV-only systems, is uplifted when GitHub is included.
However, this also leads to cases such as Carl Corp, where proprietary data is penalised. This presents a core trade-off that pillar 2 presents,
in a system where we reward demonstrated evidence, we penalise candidates who have no public work samples.


