\section{Societal Impact}
\label{sec:societal-impact}

MERIT changes the perception of ``good evidence'', and its namesake, MERIT, was an intentional reference to the concept of meritocracy.
While this may be considered a positive outcome in some cases, it is important to note that it is a trade-off that is inherent in the design of MERIT.

Relying less on CV polish and ATS-friendly formatting can help prevent candidates from being penalised for not having a perfect CV.
However, it may also lead to a shift from keyword stuffing cvs, instead towards gaming an online portfolio. Candidates could introduce
commit streaks, bloat repositories, and purposely bloating their activity metrics to appear more active. Alongside this, candidates who
work primarily in private repositories (such as enterprise, finance and defence) can score lower than juniors with active public GitHub profiles,
despite a clear difference in tenure. This is a trade-off that is inherent in the design of MERIT.

MERIT does not `fix hiring' on its own. Fair outcomes depend on the employers keeping the human veto, they must not force GitHub portfolios
as part of an application process. They must also monitor who gets rejected, and offer fair options for candidates without public work in 
order to ensure that the system is fair.
Table~\ref{tab:ethics-deployment-summary} summarises the main risks, what MERIT does about them, and what problems remain.

\begin{table}[htbp]
    \centering
    \caption{Deployment risks, MERIT mitigations, and what problems
        remain.}
    \label{tab:ethics-deployment-summary}
    \small
    \renewcommand{\arraystretch}{1.15}
    \begin{tabularx}{\textwidth}{@{}%
        >{\raggedright\arraybackslash\hsize=0.95\hsize}X
        >{\raggedright\arraybackslash\hsize=1.15\hsize}X
        >{\raggedright\arraybackslash\hsize=0.9\hsize}X @{}}
        \hline
        \textbf{Risk} & \textbf{What MERIT does} &
            \textbf{What remains} \\
        \hline
        Rejecting people with no human review (Art.~22) &
        Scores support recruiters, shortlist needs a person
            (Sec.~\ref{sec:fairness-computational-rigour}) &
        Illegal if fully automated \\
        \hline
        Bias from names, schools, titles &
        Blind Mode, smaller prestige gap in Study~15 &
        Scores still shown, tier labels kept \\
        \hline
        Users do not trust the system &
        Glass-Box scores and Shapley breakdown &
        Over-trusting scores, detail unused \\
        \hline
        Keyword stuffing on CVs &
        Anti-stuffing, multiple data sources &
        Gaming GitHub activity (Study~06) \\
        \hline
        Fake profiles / stolen GitHub &
        Integrity checks, Study~14 caught fraud &
        Rare smart squatting, name mismatches \\
        \hline
        Penalising private-sector work &
        GitHub optional, CV still used &
        ``Proprietary gap'' (Carl Corp) \\
        \hline
        LinkedIn scraping (prototype only) &
        Apify, $\leq$50 consenting contacts
            (Sec.~\ref{sec:data-privacy-gdpr}) &
        Need licensed API in production \\
        \hline
    \end{tabularx}
\end{table}
