In recent years, technical recruitment has seen a significant increase in terms of applicant volumes.
Due to this rise in the number of applications, the cognitive demands of finding a suitable candidate have
increased. This has led to more cognitively demanding Curriculum Vitae (CV) reviews, as well as the adoption of automated screening
tools to process a large volume of individuals. The current state of the art of automated screening is rigid,
black-box and also dependent on the CV as the sole source of information. This report presents \textbf{M.E.R.I.T}
(Multi-source Employment Ranking \& Insight Tool), referred to as MERIT throughout the rest of this paper.
MERIT is a prototype screening framework that supplements a candidate's CV with evidence from multiple sources,
provides flexible metrics based on the Job Description (JD), and also provides glass-box explanations for a candidate's ranking.

MERIT's design is split into three core concepts, tying into the three stages of a screening pipeline.
The first core concept is \emph{extensible} metric extraction, aiming to allow recruiters to list their own 
metrics based off the JD, targeting the ingestion stage of the screening pipeline. The next stage we target is
the scoring process itself, where we explore \emph{multi-source} fusion, combining multiple sources of information
to provide a detailed view of a candidate. Finally, once we have computed match scores for each candidate,
we provide \emph{explainable} scoring, to give more transparent and interpretable explanations for why a 
candidate was given a certain rank by our system.

The results demonstrate that multi-source fusion has the capacity to recover qualified candidates who may
have been overlooked by CV-only baselines when public work samples strongly support claimed skills. This is evidenced
by the highest observed expert rank correlation in one high-discrimination cohort (Study~07, $n=10$, Spearman $\rho = 0.697$, $p < 0.05$).
Results also point towards the effectiveness of explainability, with a formative glass-box pilot ($N=5$) indicating recruiters overturn
misleading shortlists when weak evidence or dishonest behaviour is revealed. Through our evaluation, we also observe the limitations of MERIT,
specifically parsing errors and noise, stemming from the unstructured nature of CVs and JDs. We also uncover how multi-source fusion often
penalises empty data sources, and can also deviate from expert orderings under adversarial multi-source conflict.

Therefore, MERIT is best characterised as an \textbf{auditable screening assistant} 
for human-in-the-loop review, rather than a fairness-certified or autonomous hiring authority.


