\chapter{Conclusion}
\label{ch:conclusion}

While Chapter~\ref{ch:reflection} interprets the evaluation in depth, 
this chapter states the final verdict only.

The initial question asked was whether extensible, multi-source and explainable screening systems could support
fairer, more reliable technical screenings compared to single-source rigid black-box alternatives (including typical commercial ATS rankings). 
The motivation was to move beyond CV-only screening, with its parsing fragility, self-report bias and opaque
automated rankings. Instead we pivot towards demonstrated evidence and recruiter-visible scoring. MERIT achieved
this as a working end-to-end prototype (from ingestion, to scoring, to explainability) and as an evaluation 
program of fifteen traceable studies~\cite{merit_repo}. Given the evidence in Chapter~\ref{ch:evaluation}, the answer is a 
qualified yes for research-prototype feasibility, however, a clear no for autonomous production deployment, as
mentioned in Section~\ref{sec:future-work-roadmap}.

To anchor this judgement, consider the high-discrimination cohort of Study~07 ($n=10$), where MERIT achieved the highest observed
expert rank agreement ($\rho = 0.697$, $p = 0.025$) and recovered candidates whom CV-only 
baselines discarded when GitHub demonstrated role-relevant work (Jordan Smith, Study~01B). However, we should
still note that under deliberate CV--GitHub dissonance, MERIT goes strongly against the expert consensus (Study~10).
MERIT, therefore, changes who is surfaced in favourable conditions; it does not yet guarantee correct ranking in all conditions.

MERIT is best used as an \textbf{auditable screening assistant}, in which recruiters inspect per-source evidence, treat
ranks as hypotheses, and override the engine's decisions where necessary. It may fail when treated as an oracle,
especially under conflicting sources, missing GitHub history, and role contexts that the metric model does not encode.


Against the three pillars:

\begin{itemize}
  \item \textbf{Extensible---yes, with caveats.} JD-driven metric suggestion and recruiter-overridable weights
  are implemented and partially validated (Studies~04, 05, 13). Ingestion error can still propagate into the
  rubric; configurability is the strength, document understanding the weakness.
  \item \textbf{Multi-source---yes, conditionally.} Fusion recovers hidden talent and penalises several fraud
  patterns when public evidence exists (Studies~01B, 07, 14). It also introduces open-source visibility bias,
  weak handling of empty repositories, and rank collapse under dissonance. Evidence-first ranking is the
  strength; missing-data and conflict policy the weakness.
  \item \textbf{Explainable---yes, but not sufficient for trust.} Shapley attributions, audit trails, and Blind
  Mode are implemented and tested (Studies~11, 14--15), yet transparent scores can anchor automation bias, and
  exact explainability does not scale cheaply past a few sources (Study~02). Contestability is the strength, however, 
  cost, as well as incomplete de-biasing serve as weaknesses.
\end{itemize}

\textbf{Deployment.} MERIT should not auto-reject candidates, claim fairness certification, or assert improved
hire quality without outcome-linked field studies (Section~\ref{subsec:future-validation}). Those boundaries are
developed in Chapter~\ref{ch:reflection}.

\textbf{Contribution.} 
MERIT demonstrates a coherent, evaluated alternative to the current CV-only rigid black-box ATS design. 
We document failures as well as successes, and give recruiters evidence to disagree with the machine.
It is critical to also state what was \emph{not} shown: parser $F_1$, Spearman~$\rho$, and usability scores
are feasibility indicators. They cannot serve as proof of improved hire quality, as they are not outcome-linked.

Taken together, MERIT meets its FYP aim of designing and validating a multimodal screening architecture. 
It does not yet meet the bar for unsupervised deployment in live hiring. That is a worthwhile foundation for
human-in-the-loop technical recruitment, provided the limitations in Chapter~\ref{ch:reflection} are treated as
release blockers, not footnotes. Critical reflection and a prioritised roadmap appear in
Chapter~\ref{ch:reflection}.
