\definecolor{ImperialBlue}{RGB}{0, 0, 255}    % Imperial Blue
\definecolor{ProcessBlue}{RGB}{0, 145, 212}   % Process Blue for contrast
\definecolor{LightBlue}{RGB}{235, 245, 255}   % Adjusted background blue
\definecolor{CoolGrey}{RGB}{158, 161, 162}    % Neutral Grey

\section{High-Level System Design}
The architecture of MERIT is designed to address the "Black-Box" nature of traditional
recruitment systems. It is partitioned into three core pipeline stages:
\textbf{Data Extraction and Job Configuration}, \textbf{Multi-Source Evidence Fusion}, and
\textbf{Explainable Interpretation}. Figure \ref{fig:high_level_arch} illustrates how
the recruiter interacts with the three stages of MERIT.

% Figure 4.1: High-Level System Design
\input{implementation/high_level_system_design/system_design_diagram.tex}

\subsection{Rationale}
MERIT employs a decoupled microservice architecture to separate the
concerns of data ingestion, job configuration, and the presentation of results.
We used Flask (Python) as a backend for data ingestion and fusion due to its rich ecosystem
of python libraries such as NLTK, Spacy, and Scikit-learn.
To ensure fast, prototyping and development, we used TypeScript for the frontend, 
with React and Next.js for the UI. Note that our choice of tech stack is not mandatory, it is 
just a recommendation for fast, scalable and maintainable development.

A relational PostgreSQL store (via Supabase) was selected to enforce data integrity,
ensuring that every ranked result is linked via foreign keys to a specific,
immutable recruiter configuration.

\subsection{Reliability \& Scalability}
While prototyping MERIT, we encounntered many major obstacles that provided 
significant obstacles to the development of the system. Below, are the two 
most significant challenges that we faced:

\subsubsection{The N+1 Query Bottleneck}
A significant scalability challenge was the \textbf{N+1 query bottleneck}. When
retrieving $N$ candidates, the backend was previously making $N$ subsequent database
calls to fetch associated skills and scores. By leveraging PostgreSQL's relational
embedding capabilities via the Supabase client, we transitioned query complexity from
$O(N)$ to $O(1)$. The benefits of this runtime optimisation are significant when considering
a large candidate pool, which will be almost always the case in real-world
applications.

\subsubsection{Global Error Handling}
Due to the lack of type safety in python, we encountered numerous runtime errors throughout prototyping.
One of the most significant challenges we faces was in the parsing of unstructured data (CVs and JDs).
Our approach to overcome this was in the form of the introduction of global Flask middleware to
intercept any exceptions throughout the backend pipeline. Followiung this up with strict JSON schema validation ensures
that only verified signals enter the multi-source fusion engine. This helped ensure the integrity of the explainability layer.
