\section{Stage 1: Data Extraction and Job Configuration}
The first stage of the MERIT pipeline focuses on \textit{Data Extraction} and \textit{Job Configuration}.
Here, we translate unstructured job requirements into a structured, weighted mathematically-backed profile.
This stage is critical for establishing the system's "Integrity Signal". Rather than requiring
recruiters to manually define every skill and priority, MERIT leverages automated extraction
and market-aware weighting to pre-fill the job's configuration.
This process, illustrated in Figure \ref{fig:skill_extraction_pipeline}, ensures that the system remains
extensible to any technical domain while significantly reducing the manual overhead of
job setup.


\input{implementation/high_level_system_design/skill_extraction_diagram.tex}

\subsection{Ingestion \& Normalisation}
The first step in this stage is the ingestion and normalisation of heterogeneous data sources.
As illustrated in Figure \ref{fig:extraction_flow}, MERIT handles both unstructured documents
(CVs and JDs) and structured API signals (GitHub and LinkedIn). This layer performs
canonical skill mapping to ensure that signals from different sources can be compared
within a unified schema.

\input{implementation/high_level_system_design/extraction_layer_diagram.tex}

\subsection{Skill Extraction \& Preprocessing}
Before we perform semantic analysis, we first need to extract the relevant technical
skills from the job descriptions and CVs. For MERIT, we have implemented a heuristic
entity extraction method that utilises a curated dictionary of over 500 technical
languages and frameworks (stored in \texttt{skills\_data.json}). This system identifies
raw mentions of these terms within the text. This helps us isolate technical signals from general
prose, to ensure that the semantic matcher only focuses on relevant skills. Another purpose of this
preprocessing step is to resolve common variations of the same skill. For example, the term
"JS" is often used interchangeably with "JavaScript" in job descriptions. By identifying and
normalising these variations, we prevent the inclusion of unnecessary noise during the vector similarity calculation.

\subsection{Semantic Skill Matching}
After we have performed a relatively rigid entity extraction, we move onto the
semantic matching stage. Here, we address the limitations of traditional Applicant
Tracking Systems (ATS) that rely on exact string matching. We are able to achieve this through
the use of Latent Space Mathematics. Rather than following traditional ATS systems that rely on
exact string matching, we instead rely on vector-based similarity analysis.

We use the \textit{SentenceTransformers} library, and the \textit{all-MiniLM-L6-v2} model for this process.
This model was selected for its balance between efficiency and accuracy for technical terminology.
We encode terms into a 384-dimensional dense vector space, capturing the underlying semantic relationships
between technologies. This is achieved through the application of the \textbf{cosine similarity}
formula as demonstrated below.

\begin{equation}
    \text{similarity} = \cos(\theta) = \frac{\mathbf{A} \cdot \mathbf{B}}{\|\mathbf{A}\| \|\mathbf{B}\|}
\end{equation}

For MERIT, we set the threshold to $0.50$, this was determined through iterative testing and
comparison with the results of the heuristic entity extraction method. This threshold is
significant enough to determine that "SQL" and "PostgreSQL" are similar skills, while not being
too strict as to reject semantically similar terms such as "React" and "React.js"


\subsection{Dynamic Metric Weighting}
Another key innovation towards the extensibility of MERIT lies in the inference of skill importance
when setting weights during the configuration stage. This is achieved through the application of a
modified Term Frequency-Inverse Document Frequency (TF-IDF) approach. This aligns with foundational
Information Retrieval theory \cite{sparckjones1972idf}, which establishes that skills with a high
Inverse Document Frequency (IDF) are significantly more "informative" for distinguishing specific
job roles than universally common terms.

We calculate the TF-IDF scores based on both the job description, as well as the market corpus.
The \textbf{Term Frequency (TF)}, or "Job Intensity," reflects the employer's subjective emphasis
and is calculated as:

\begin{equation}
    TF(s, J) = \frac{f_{s, J}}{\sum_{k \in J} w_k}
\end{equation}

Where $f_{s, J}$ is the number of times skill $s$ is mentioned in the job description
$J$, and $\sum w_k$ is the total word count of the document. Simultaneously, the \textbf{Inverse Document Frequency (IDF)},
or "Market Scarcity," is calculated as:

\begin{equation}
    IDF(s, D) = \log\left(\frac{N}{1 + df(s)}\right) + 1
\end{equation}

Where $N$ is the total number of job descriptions in our market corpus $D$, and $df(s)$ is the
number of documents containing skill $s$ \cite{bastian2014linkedin}. The final significance score
$W_s$ for each skill is the product of these two components:

\begin{equation}
    W_s = TF(s, J) \times IDF(s, D)
\end{equation}

For each configured job, we compute the $W_s$ scores for each skill, this may also commonly referred to as an extensible metric.
These raw scores are the product of the TF and IDF scores for each skill. From here, we then min-max normalise the scores
to the range $[1.0, 5.0]$ with one decimal place. Higher values mean stronger emphasis on that skill for the given job.
If a JD mentions both a common skill (e.g.\ ``Python'') and a rarer one (e.g.\ ``Jupyter''), the rarer term tends toward 
the top of the band unless the common skill is repeatedly mentioned in the JD. The market corpus $D$ is populated from
historical job descriptions held in Supabase (or alternatively could be populated from an offline corpus, such as during evaluation).
Note that the performance of TF-IDF is dependent on the market corpus, and as such, if the market corpus is not representative of 
the current market, the performance of TF-IDF will be affected. This allows MERIT to keep up to date with the current market, and to be 
extensible to any technical domain.


\subsection{Target Metric Weights}
Weights should never be fixed, TF-IDF is merely a suggestion based off the market corpus. As such, we also allow recruiters
to manually adjust the weights of the skills on the job. This is achieved through the use of a 5-point Likert scale, where
1 = essential and 5 = almost optional. This choice is supported by psychometric research on rating-scale length 
\cite{lozano2008effect, coelho2007choice}. Note it is possible to completely disregard a metric as well if the recruiter does 
not wish to include that metric in the candidate matching process. Note weightings in the backend and frontend are inverses
due to the way that Likert scales correlate to the target weights, as such, in the backend, higher weights mean more significant
for that skill on that job, while in the frontend, lower weights mean more significant. Simply put, the frontend weights are given
by the formula:
\begin{equation}
    P_{\text{UI}} = \max\left(1,\; \min\left(5,\; \mathrm{round}(6 - \texttt{suggested\_weight})\right)\right).
\end{equation}
Thus the strongest skill on a JD ($5.0$) is shown as P-1 and the weakest ($1.0$) as P-5.
