\section{Stage 2: Multi-Source Candidate Ranking}
Once candidate data and a job description have been extracted and combined in a configuration, we can then proceed to
the execution stage. In this stage, we compute signals for each metric. This can be done in two different ways
depending on whether the metric is baseline or extensible. Both metrics may take into account multiple data sources,
computing up to one signal per data source per metric. If this is the case, we perform Bayesian fusion to compute a single
score. Once each metric has been reduced to a single value, we then undergo a weighted average of all the metrics
to compute the final (match) score. Figure \ref{fig:schema_overview} shows how we can use a candidate's
profile alongside a job description to produce a final score: candidates from multiple sources are grouped into a batch,
paired with an extracted job description to form a configuration, and each configuration run stores an immutable past result.


\input{implementation/high_level_system_design/relational_schema_diagram.tex}

\subsection{Baseline vs. Extensible Metrics}
When executing a job configuration, the system, categorises evaluation criteria against two different kinds of metrics.
\textbf{Baseline Metrics} are pre-defined metrics that are common to all configurations, and typically follow the industry-standard
heuristics. For example, a candidate's education level is a universal factor in technical recruitment that can be assessed
independently of the specific job requirements. \textbf{Extensible Metrics}, on the other hand, are job-specific metrics, in other words,
they are directly inferred from the Job Description. An example of an extensible metric for a typical "Machine Learning Engineer"
may be "Python", which would ideally measure the candidate's skill level, or usage of the programming language. By combining these
two classes of metrics, we are able to produce a holistic, yet highly targeted evaluation of a candidate.

For both classes of metrics, the system may take into account multiple data sources. Some of these sources may be self-reported
(such as CV and LinkedIn) while other sources may be evidence-driven (such as GitHub). After reducing these signals to a single score,
we perform a weighted average of all the metrics to compute the final (match) score.

\subsection{Semantic Skill Matching}
\label{subsec:semantic-skill-matching}

One of the biggest issues with keyword-matching screening systems lies in their rigidity.
Suppose a candidate applied to a role where ``SQL'' is required, however, the candidate only uses
``PostgreSQL'' or ``Postgres'' in their CV. This candidate would likely achieve a zero for the ``SQL'' metric.
This is where semantic skill matching (or bridging) comes in. When there are no exact matches,
we perform a cosine similarity operation, using the \textit{SentenceTransformers} library with the 
\textit{all-MiniLM-L6-v2} model. We selected this model due to its efficiency and accuracy regarding
technical terminology. 

For this model, we encode the skills into a 384-dimensional vector space, and then compute cosine
similarity.

\begin{equation}
    \text{similarity} = \cos(\theta) = \frac{\mathbf{A} \cdot \mathbf{B}}{\|\mathbf{A}\| \|\mathbf{B}\|}
\end{equation}

To be considered as a semantic bridge, there must be a cosine similarity score of $\geq 0.50$ between
the extensible metric and the skill mentioned in the candidate CV. This threshold was determined 
through iterative testing. The threshold is permissive enough to match terms such as "SQL" and "PostgreSQL",
while also remaining strict enough to reject other, unrelated terms such as ``Python'' and ``JavaScript''.
Whenever there is a semantic bridge, we also pass this information to the frontend to let the recruiter know
of both extensible metric, as well as the skill that was bridged against it, to ensure full transparency.
This way, in the event of a false-positive, the recruiter can easily identify and correct it.
For extensible CV metrics, a direct string match (e.g.\ ``SQL'') yields a lexical signal, otherwise cosine similarity
against extracted skills is used (e.g.\ bridging ``PostgreSQL'' to ``SQL''), as summarised in Figure~\ref{fig:semantic_skill_scoring}.


\input{implementation/high_level_system_design/semantic_skill_scoring_diagram.tex}

\subsection{Integrity Layer}
A common flaw in legacy ATS platforms lies in their vulnerability to adversarial attacks, such as
keyword-stuffing or identity squatting. To combat these, MERIT incorporates an Integrity Layer
that executes concurrently with the signal extraction process. While the detailed heuristics for
these safeguards are discussed in Section 4.5, it is important to note that these penalties are applied
directly to the raw evidence signals before they enter the Bayesian fusion pipeline. This ensures that
poisoned data (in other words, data that warrants a penalty) carries less significance in the final probabilistic
model if other signals are still robust. For example, a candidate who gets penalised for keyword stuffing in their CV may
still rank highly if they have strong evidence from other sources for that metric.

\subsubsection{Skill Decay Modelling}
\label{subsec:skill-decay-modelling}
Skill decay is a concept that is used to model the loss of a skill over time, given that the skill has not been used for a period of time.
Applying this to technical recruitment, it is intuitive that the recency of a candidate's skill matters. A candidate may have been considered
an expert in a certain programming language ten years ago, but if they had not used it since then, would their level be considered the same?
They may not carry the same weight as a candidate who has used it often within the past few months, and prior research shows that this skill
loss grows with longer periods of non-use \cite{arthur1998}. Industry research points to a similar finding, where workplace skills (especially technical skills)
become outdated far more quickly than they once did \cite{ibm2019skillsgap,wef2020futurejobs}. Traditional ATS platforms fail to account for this since
there is often no significant evidence to uncover when the candidate last used the skill in question.

MERIT attempts to overcome this issue through the addition of an exponential skill decay model for extensible metrics (e.g.\ ``Python'').
We apply this to the GitHub signal, as it is the most reliable signal for demonstrating evidence of a candidate's skill.
For other signals, it is usually harder to find a point to start the decay from, but for GitHub, we can determine it from the last
point of significant activity. In our backend, we get the repositories' language distribution, spread it across its active years and use that
to calculate the last point of significant activity, using that year as the start of the decay. Subtracting this value from the current
date nets us with the $\Delta t$ value in the decay formula:

\begin{equation}
  \mathcal{S}_{\text{decay}} = \mathcal{S}_{\text{base}} \cdot e^{-\lambda \Delta t}
\end{equation}

Where:
\begin{itemize}
  \item $\mathcal{S}_{\text{base}}$ is the initial magnitude of the extracted skill signal.
  \item $\Delta t$ is the time elapsed (in years) since the end date of the relevant activity.
  \item $\lambda$ is the decay constant, determining the rate of skill degradation.
\end{itemize}

The system is configured with a default exponential decay constant of $\lambda = 0.25$,
which corresponds to a half-life of $t_{1/2} = \ln(2)/\lambda \approx 2.8$ years.
This value is not taken directly from a single empirical study. Instead, it is a practical default chosen to align with
industry estimates that technical skills can lose much of their relevance within roughly 2.5--5 years \cite{ibm2019skillsgap}.
This default value ensures that recent, relevant experience naturally outranks historical,
potentially obsolete capabilities.

\input{implementation/high_level_system_design/skill_decay_graph.tex}


\subsection{Evidence-Based Signal Aggregation}
The reasoning for MERIT's multi-source design is to be able to consider different sources of candidate data,
those that are self-reported by the user, and those that the user has actually been able to demonstrate
proficiency in. MERIT splits these into two different forms of evidence, \textbf{Claimed Evidence} and 
\textbf{Demonstrated Evidence}.

\textbf{Claimed evidence} refers to a self-declared proficiency in a skill, programming language or framework.
This form of evidence is commonly found in CVs and LinkedIn profiles due to the user being responsible
for their own claims. This is the only form of evidence usually considered by ATS scoring engines, making
these systems biased towards candidates who are able to present their skills in a better light. Self-embellishment
is often never punished, as it is difficult to verify, but this is where the beauty of MERIT comes into play,
with its second form of evidence.

\textbf{Demonstrated Evidence} refers to evidence derived from auditable, historical actions. In our case, this is 
specifically GitHub repositories. For example, a candidate is likely to claim Python proficiency on their CV for 
a job application, however, this claim can be challenged by the presence of this skill on their GitHub repositories.
Demonstrated evidence is significantly harder to game as it requires a candidate to actually use it as part of 
their own projects or daily work.

This formal separation of evidence is key to MERIT's "Evidence-based evaluation" ethos, and will 
play a key role in Bayesian Fusion.

\subsection{Bayesian Fusion}
The most significant innovation in MERIT's Multi-Source Fusion module lies in the adoption of \textbf{Bayesian Evidence Fusion}.
Initial iterations of MERIT attempted to combine skill signals using static weighted averages. However, this approach often dilutes
the importance of demonstrated evidence (GitHub) with claimed evidence (CV and LinkedIn). This approach also failed to account for the
confidence of a measurement. For example, a single line in a CV stating "Python" does not carry the same weight as a GitHub profile
that demonstrates extensive Python experience.

Another key issue is the lack of flexibility it provides, as we would now need separate definitions to account for candidates who may not
have all three data sources. Several considered approaches failed to provide a fair solution whilst adhering to the static weighted average
approach. Alongside this, there was also the issue of extensibility, as we need to provide $N!$ distinct definitions for $N$ data sources if
we wish to avoid penalising candidates for missing data sources.

The presence of these dilemmas led to the adoption of \textbf{Bayesian Evidence Fusion}. The current iteration of MERIT now treats each
data source as independent evidence updating a Beta Distribution. We parameterise this with the use of a sparse prior
($\alpha_{prior} = 0.1, \beta_{prior} = 0.1$) to ensure high sensitivity to initial evidence signals while maintaining a baseline for uncertainty.

The contribution of each source $s$ to the final score of that metric is given by its raw signal $\sigma_s \in [0, 1]$ and its reliability weight $w_s$.
Evidence-based data sources are often assigned a higher weight due to MERIT's emphasis on evidence-based screening. To encourage candidates to
use evidence-based sources while avoiding harsh penalisation, we enforce a ``Signal Cap'' for sources that provide claimed evidence.
We then map these signals to the evidence parameters as follows:
\begin{equation}
  \alpha_s = w_s \sigma_s + \epsilon_s, \quad \beta_s = w_s (1 - \sigma_s) + \epsilon_s
\end{equation}
where $\epsilon_s = 0.01(1 - w_s)$ represents a symmetric uncertainty term. The reason for the inclusion of $\epsilon_s$ is to inject noise for
low-confidence sources to produce wider, more uncertain distributions. This stops the distribution from becoming narrow unless this candidate has the
evidence to support their claims (since high-confidence sources provide more concentrated signals.)

We select the Beta distribution since it is the conjugate prior for Bernoulli-distributed evidence. This means that the resulting posterior
update still remains a Beta distribution. This allows for additive evidence integration, which is, not only computationally more efficient,
but also extensible for an arbitrary set of data sources $S$. This aligns with the "Extensibility" goal of MERIT. The updates to the Beta
distribution are given by:
\begin{equation}
  \alpha_{total} = \alpha_{prior} + \sum_{s \in S} \alpha_s, \quad \beta_{total} = \beta_{prior} + \sum_{s \in S} \beta_s
\end{equation}

The final posterior distribution is:
\begin{equation}
  P(\theta \mid \text{Evidence}) \sim \text{Beta}(\alpha_{total}, \beta_{total})
\end{equation}

This probabilistic approach provides a robust framework for modelling uncertainty and resolving conflicts between
different evidence sources. For a verified signal, if a candidate has claimed evidence on a CV (low weight) and also
possesses the demonstrated evidence to back it up, such as on GitHub (high weight), then the resulting distribution will
skew towards a high probability of the candidate possessing the skill, with low variance. On the other hand, if a skill is listed
heavily on their CV, but absent from their GitHub, the $\alpha$ parameter will only increase slightly. The baseline $\beta$ parameter
will then prevent the score from artificially inflating, mathematically mitigating "keyword stuffing" even further.

Once we have computed the posterior distribution for a metric, we can then split it into its expected value and its variance. Effectively
this measures what the candidate's proficiency in a certain skill is, and the confidence that we have in that claim. For a random variable
$\theta \sim \text{Beta}(\alpha, \beta)$, we define its expected value $E[\theta]$ and variance $\text{Var}(\theta)$ as:
\begin{equation}
  E[\theta] = \frac{\alpha}{\alpha + \beta}, \quad \text{Var}(\theta) = \frac{\alpha\beta}{(\alpha + \beta)^2(\alpha + \beta + 1)}
\end{equation}

To conclude, the expected value of this distribution is the candidate's final fused score for the metric in question.
The ranking system, as a result, aims to prioritise candidates with verified evidence across multiple sources over sources
which only provide claimed evidence. However, this is not strict enough to entirely rule out candidates with claimed evidence
who may not have evidence to back it up.





\subsection{Glass-Box Payload}
The majority of modern ATS platforms receive criticism for being black-box in nature. It is often unclear to recruiters exactly
how a candidate was ranked, nor can it delve into the specific metrics that contributed to a candidate's final score. 
MERIT's attempt to address this core problem is through returning a JSON payload that functions as a mathematical audit trail.
We show the full scoring process for a given metric, from the raw data being extracted, to the temporal decay applied to
demonstrated evidence signals, to the Bayesian Fusion process. Where penalties are applied (such as keyword stuffing), they
are also raised as well. 
We also store a lightweight \texttt{summary\_data} object\footnote{The \texttt{summary\_data} field is a lightweight UI-optimisation
  feature designed to populate recruiter dashboards without the network latency of fetching the complete, multi-megabyte
  \texttt{results\_payload}.}, which contains high-level
metadata (such as the top candidate and score) for rapid batch retrieval without
re-processing the full evaluation payload.

The payload is a final product of the execution stage. Its structure ensures that every high-level score is
backed by the granular source-specific evidence that it's comprised of. The \texttt{results\_payload} schema
consists of three main layers, as follows.

The first layer is the highest level layer, the aggregation layer, which contains the overall score and the
calculation summary, providing a high-level overview and the global weighted
formula used. The second layer is the middle layer, the metric layer. This contains the specific evaluation criteria
(e.g., \textit{Experience}, \textit{req\_Python}) stated in the job configuration. Each entry contains its independent
Bayesian parameters ($\alpha, \beta$) and its fused score where necessary. The third layer is the evidence layer,
which is the lowest level layer, found within \texttt{source\_details}. This contains the specific
string derivations, source confidence coefficients, and temporal decay penalties applied to individual signals.

By exposing this structured data in a candidate's report, we justify the score for a given candidate: recruiters
can see how each data source contributed for a given metric, how Bayesian fusion combined those contributions into
a per-metric score, and how metrics are weighted by job priority to produce the final match score. Presenting every
stage in this form gives recruiters the means to challenge and interrogate the score rather than accepting it as
a black box. Appendix~\ref{appendix:math_trail} shows a representative end-to-end audit trail in the UI.
