\section{Automated Screening}
\label{sec:automated-screening}

Automated screening systems (commonly referred to as Applicant Tracking Systems (ATS)), are used by many organisations to filter, rank and manage large volumes of candidates.
Among large firms, usage is common, with Jobscan industry reports on Fortune 500 ATS adoption consistently finding rates above 97\% across survey editions \cite{jobscan2022}.
While ATS adoption rates outside of the largest firms are not as certain, these systems remain a core part of scaling the recruitment pipeline in high-volume hiring environments.

While ATS implementations differ significantly, they can be categorised into one of three types: Rule-Based, Keyword-Matching, and AI-Assisted.

\subsection{Rule-Based ATS}
These forms of ATS were the first to be introduced, and often the simplest.
They work by parsing resumes into structured fields, such as skills, education, job titles and dates.
From here, explicit, pre-defined rules are applied in order to screen candidates.
Recruiters often configure keyword lists, required skill thresholds, boolean queries and knockout conditions (e.g., minimum degree requirements and years of experience).

Due to their consistency and predictability, employers may leverage rule-based systems to screen candidates. 
By applying explicit, pre-defined criteria uniformly, such systems avoid the intuitive, subjective judgments that 
often influence human reviewers \cite{highhouse2008stubborn}.

The reason these systems are not commonly used however, is due to the limit of standardisation of CV formats.
These systems will struggle with non-standard CV formats, leading to parsing errors \cite{emnlp2025resumebench}.
Candidates with unconventional layouts or career paths are therefore more likely to be scored poorly for formatting 
reasons rather than skill.
Along with this, these systems are vulnerable to keyword-gaming, a practice where applicants strategically insert 
repeated or irrelevant terms to artificially inflate their match scores, a practice that is widely discussed in practitioner advice on optimising CVs for ATS parsers \cite{marr2019}.
The failure to recognise all possible CV formats and lack of semantic understanding also leads to the system being unable
to recognise unconventional experience. It may also perform poorly when skills are also presented in non-traditional ways, or
demonstrated in another section, such as open-source contributions or bootcamp projects. Rule-based systems are likely not 
exhaustive enough in the treatment of CVs and will likely score non-traditional CVs poorly, not due to the candidate's skills,
but instead due to the system's inability to recognise the candidate's skills.


\subsection{Keyword-Matching ATS}
This version of ATS is a middle ground between the previously mentioned rule-based approach, and the upcoming AI-driven version.
This form still relied on keyword matching, but also incorporates better NLP-based parsing to recognise linguistic variations and identify skills even when phrased differently (e.g., “managed projects” → “project management”).
These systems offer limited semantic understanding, able to recognise that related job titles or synonyms carry similar meanings.

Their strengths are mainly that there is increased flexibility and better extraction accuracy.
There is also the improved ability in matching candidates who used varied phrasing, and also that there is more transparency than the AI-driven systems, since much of the logic is still rule-based.

Despite the improvements, many issues still persist, especially with interpretation.
These systems usually misinterpret contextual details as well as ambiguous language within a resume.
The lack of sophistication also may lead to them overlooking crucial nuances such as depth of a skill, the quality of a candidate’s experience, as well as specific evidence of measurable impact.
Due to this, the limitations often result in misclassification: employer surveys report that more than 90\% use applicant-tracking or recruitment-management systems to filter or rank candidates, and 88--94\% of employers say qualified applicants are screened out when they fail to match job-description criteria exactly \cite{fuller2021hiddenworkers}.

\subsection{AI-Assisted ATS}
This form of ATS leverages machine learning models that have been trained on historical hiring data to predict candidate-job fit.
Unlike the previous two groups AI-assisted ATS don't rely on keyword matching and  instead infer relationships between skills, roles, qualifications and past hiring decisions.

AI-Assisted ATS usually use semantic search and vector-based skill matchmaking to understand the conceptual intent behind a resume, alongside skill inference drawn from conceptual data.
They also employ predictive scoring models that learn from recruiter interactions, and utilise pattern recognition to identify hidden, non-obvious candidate strengths that may otherwise go unnoticed.

Their strengths include higher matchmaking accuracy, as well as the ability to detect transferable skills, not explicitly written on CVs. They also handle non-traditional career paths more effectively than rule-based systems.

However, there are also many challenges faced by AI-Assisted ATS systems too.
Firstly, algorithmic bias is a notable issue where models inherit and reproduce historical patterns of gender, ethnic or institutional preference found in past hiring data \cite{raghavan2020,bogen2018}.
This issue is made worse with feedback loops, as systems that consistently prioritise specific candidate profiles may continue to further these traits even as a company's strategic needs change, as noted in reviews of algorithmic hiring \cite{raghavan2020}.

Furthermore, the “black box” nature of these systems lead to low transparency, as there is often little explanation for ranking and rejecting decisions, undermining both the candidate trust as well as compliance with fair-hiring guidelines.
Another significant issue lies in over-generalisation, where these models are often trained on broad datasets, prioritising generic markers over role-specific expertise.
In a similar fashion, domain drift also plays a role, where the system continues to optimise for outdated historical traits despite a company's technological or strategic shifts.

It is critical to note that almost all ATS platforms operate exclusively on CV and application data, limiting their ability to capture verified evidence of a candidate's true capabilities.
MERIT addresses this directly by introducing a multimodal evidence-driven screening approach that integrates behavioural, technical and performance-based data from various sources.
If done successfully, MERIT positions itself as a prototype framework that combines multimodal candidate profiling with transparent, explainable AI, two capabilities largely absent from existing systems.
This enables recruiters to make decisions grounded in observable performance rather than self-reported claims, something that neither traditional nor modern ATS have been able to achieve.