\section{Data Sources}
\label{sec:data-sources}
Technical recruitment often requires multiple data sources to be provided to evaluate candidates beyond traditional CVs, but are often rarely used. These sources have the potential to provide insights into the extensive professional history, technical competences, behavioural tendencies as well as real-world coding practices. Effective multi-sources screening will require a deeper understanding on the strengths and limitations of these data channels.

\subsection{Commonly Provided data sources}
The following data sources are commonly provided by candidates during technical recruitment processes:

\subsubsection*{Curriculum Vitae}
A CV remains the primary go-to document for recruitment pipelines, for technical roles it is especially important, allowing recruiters to quickly compare candidates without the need to commit to an extensive interview cycle for every candidate. 
A typical technical CV will include a candidate's name, education, employment history, projects and technical skills. 
Despite the amount of key data provided, research shows that CVs can contain unreliable self-representations due to intentional embellishment, omission, or fabrication \cite{henle2019_resume_fraud,dunning2004}. 
For this very reason, organisations increasingly supplement CVs with behavioural and performance-based evidence.

\subsubsection*{GitHub} 
GitHub is a central platform for hosting projects and version-controlled repositories. 
For developers, it is arguably the richest source of work-sample evidence, capturing coding style, commit history, collaboration practices, documentation depth, and long-term engagement patterns. 
Industry research indicates that portfolio evidence such as GitHub is widely valued in technical hiring. In HackerRank's 2018 survey, 80\% of small companies and 66\% of large companies reported placing higher importance on portfolios \cite{hackerrank2018skills}. However, review of these repositories by recruiters is often inconsistent, and may be limited to high-level elements such as README files rather than deeper technical behaviour. 
Despite this, GitHub remains a highly valuable source of work-sample evidence.

\subsubsection*{LinkedIn} LinkedIn functions as a professional identity platform, it allows users to expand upon their projects, experiences and education, offering significantly more information that what may be provided on a CV. 
It also functions as a professional social media platform, where users can share their opinions on different topics. 
LinkedIn is widely used in recruiter workflows; in Jobvite's 2020 Recruiter Nation survey, 72\% of recruiters reported using or planning to use LinkedIn for recruitment \cite{jobvite2020}.

\subsection{Other data sources}
There are also other data sources that, while not often provided, can offer valuable insights into a candidate's suitability for a technical role. These include:


\subsubsection*{Leetcode} LeetCode is one of the most widely used platforms for practicing algorithmic problems that often appear in technical interviews and online assessments. 
Industry surveys show that structured coding assessments are common in technical hiring: in CoderPad's 2025 global survey, 49.8\% of employers reported using live coding interviews and 24.5\% reported using asynchronous technical tests, many of which focus on algorithmic problems similar to those on LeetCode \cite{coderpad2025techhiring}. 
Because of this, a candidate's activity on coding platforms can reveal problem-solving patterns, topic strengths, and recent engagement. It may also highlight areas where candidates are weaker, helping recruiters assess adaptability to new or unfamiliar problems.

\subsubsection*{HackerRank} HackerRank is quite similar to LeetCode, in terms of it being a platform for coding challenges. 
It is also widely used as a platform to host online technical tests, setting it apart from its competitor. 
Many companies, especially those hiring at scale rely on HackerRank for standardised assessments, making the platform an interesting data source for understanding a candidate's tested technical ability.

\subsubsection*{Stack Overflow} Stack Overflow is a Q\&A platform for software engineering knowledge exchange. 
Contributions on Stack Overflow can provide insight into a candidate's communication style, community engagement, and areas of expertise. 
Prior research shows that contributions on such platforms can correlate with professional experience and also domain specialisation \cite{vasilescu2013}.

\subsubsection*{Social Media} During background checks, organisations may review social flags, though this is often to identify red flags rather than evaluate technical competence. 
Research shows that 70\% of employers use social media to research job candidates during hiring \cite{careerbuilder2017_social}. 
Due to the practice being controversial due to privacy, bias and fairness concerns, MERIT will not consider non-professional social media as a supplementary data source.
Note that professional social media such as LinkedIn is still considered.

\subsection*{Relevance to MERIT}
As shown above, there are a variety of external data sources that can be used to evaluate a candidate's suitability for a technical role
beyond the traditional Curriculum Vitae (CV). Being able to capture and integrate multiple data sources can help provide more factors
to differentiate candidates between, allowing for a more comprehensive view of a candidate's suitability for a role. MERIT aims to 
integrate GitHub as a supplementary data source due to it serving as a strong indicator of a candidate's technical ability. LinkedIn
will also be selected as a supplementary data source due to it serving as a strong indicator of a candidate's professional experience.
Other data sources such as LeetCode and HackerRank are likely to not be selected as supplementary data sources due to the time 
constraints of this project, however, they are still considered as potential future data sources for MERIT.


