\section{Job Description}

A job description (JD) serves as a formal specification of a role’s purpose and scope, often starting off with a clear definition of the position’s objectives as well as its placement within the organisation’s hierarchy. 
For technical roles, essential responsibilities such as programming, deployment and research may also be listed, as well as necessary technical competencies in specific languages, frameworks and tools. 
Alongside this, JDs often point out desired soft skills such as communication and collaboration. 
Organisational context, team structure, culture and growth opportunities may also be provided to give a candidate a better insight into the role.
\\\\
However, due to the lack of an industry standard for how JDs are written, they will often vary in clarity, completeness and realism. 
While some JDs are detailed and explicit, others are vague and overly generic. 
According to research on realistic job previews (RJPs), when JDs or job advertisement lack accuracy and a balanced picture of job demands, there may be unrealistic expectations as a result, especially when negative aspects are omitted, or benefits are inflates. 
This imperfect information often results in dissatisfaction or turnover once hired \cite{breaugh1983}. Similarly, empirical evidence shows that a more realistic, transparent JD can correlate with a better candidate-job fit, as well as a lower voluntary turnover \cite{earnest2011}.
\\\\
Given the complexities of technical roles, a rigid JD may fail to accurately reflect all dimensions of the role. 
This increases the risk that initial screenings (regardless of whether manual or automated) may miss important signals, or over screens based on superficial or inessential attributes.
\\\\
Therefore, MERIT must treat JDs as flexible, interpretive inputs, extracting both explicit requirements as well as implied/behavioural needs and mapping these against multimodal candidate evidence. 
This will allow for a more accurate evaluation of fit, reducing dependence on self-reported credentials and also mitigates the risks associated with an oversimplified or inflated job description.