\section{Summary of Gaps and Relevance to MERIT}
Based on the review of current state-of-the-art research and implementation of 
automated technical screening systems, three fundamental gaps can be identified. 
These limitations are the direct motivators for the design and implementation of MERIT.


\subsection*{The Extensibility Gap}
Existing platforms often operate on static, hardcoded role classifications and fixed scoring dimensions. 
As noted by Gugnani and Misra \cite{gugnani2020}, rigid scoring systems hinder the 
ability to adapt to the nuanced requirements of diverse engineering roles. 
Current systems are unable to adapt to the ever evolving needs of the tech industry.
The relevance of a given langauge or framework can change drastically from year to year, and often, these
systems are unable to adapt based on the changing perception of the recruiter market. 
As a result, there is a need for a system that is able to dynamically adapt its evaluation criteria based 
on, not only the job description, but also the fluctuating trends in the technical recruitment industry.


\subsection*{The Multi-Source Verification Gap}
The majority of ATS, even those built upon modern LLMs remain dependent on the self-reported claims within a CV.
Research shows that these documents can contain unreliable self-representations due to intentional embellishment, omission, 
or fabrication \cite{henle2019_resume_fraud,dunning2004}. Due to this single-source dependency, screening often relies on claims 
that are difficult to verify, while candidates may also tailor their CVs to match the keywords in a job description in order to 
pass through ATS filters \cite{fuller2021hiddenworkers}. While there is research on the potential of using external signals
from sources such as GitHub \cite{montandon2021_roles}, existing approaches mainly infer coarse technical-role labels from repository metadata,
rather than verifying CV claims with traceable, multi-source evidence of competence. Due to this, there is a need for a system that can
cross-reference CV claims against observable evidence from external data sources that are harder to forge and embellish.
This system must be able to resolve conflicts between these sources to establish a verifiable "ground truth" for a given 
candidate's competence.


\subsection*{The Transparency and Interpretability Gap}
Innovations in NLP have allowed for greater accuracy in candidate screening systems, however, this 
often comes at the cost of transparency and interpretability. This has introduced a "transparency deficit" \cite{bogen2018}, 
where recruiters are presented with scores that lack granular justification, making it difficult to interpret why 
a candidate was given a particular score. The lack of auditability complicates compliance with fair-hiring regulations \cite{raghavan2020,bogen2018}.
To address this, we will require a shift towards explainable architectures that can provide an audit trail for decisions, allowing
recruiters to understand the reasoning behind a candidate's score.


\subsection*{The MERIT Approach}
MERIT aims to tackle these limitations through three core innovations. Firstly, it introduces a 
flexible role classification system that can be easily adapted to the evolving needs of the technical
recruitment industry. Secondly, it incorporates data from multiple sources, including CV, GitHub, 
and LinkedIn, to provide a comprehensive view of a candidate's skills and experiences. Thirdly, it 
provides explainable outputs that allow recruiters to understand the reasoning behind a candidate's 
score. By doing so, MERIT introduces a new standard for verifiable, transparent and context-sensitive
candidate screening, and the implementation of which, will be disclosed in the following chapter.
