\section{Semantic Intelligence and Vector Embeddings}

MERIT leverages the transition from traditional lexical analysis to semantic intelligence to evaluate the technical expertise of a candidate.
Earlier generations of ATS relied on "keyword matching", however, MERIT will leverage Large Language Models (LLMs) and vector embeddings to build relationships between programming languages and technical frameworks.

\subsection*{Lexical vs. Semantic Analysis}
Traditional lexical methods operate on the principle of exact character matching. 
These methods are often efficient for identifying rare terms, such as a unique framework.
However, they are vulnerable to the "vocabulary mismatch" problem, where a candidate may not use the exact same terminology as the job description.
Failing to match a candidate due to a mismatch in terminology is a significant issue that can lead to qualified candidates being overlooked simply because they didn't use the same keywords as the job description.

Semantic models, on the other hand, represent technical skills as vectors in a high-dimensional latent space. 
By mapping technologies to coordinates, MERIT can measure the proximity of skills rather than the frequency of words. 
This allows the system to understand that these terms are conceptually related, enabling "soft" matching for candidates who possess highly transferable technical skills.

\subsection*{Vector Similarity Measures}
The similarity between a candidate's technical profile ($\mathbf{C}$) and the job requirements ($\mathbf{J}$) is calculated using measures such as Cosine Similarity:
\begin{equation}
    \text{similarity} = \cos(\theta) = \frac{\mathbf{C} \cdot \mathbf{J}}{\|\mathbf{C}\| \|\mathbf{J}\|}
\end{equation}

This mathematical foundation enables MERIT to identify related competencies and technical potential that would be missed by rigid rule-based systems. 
By utilising transformer-based architectures such as Sentence-BERT \cite{reimers2019sentence}, the system can generate dense vector representations that capture the subtle linguistic nuances of technical expertise and framework relationships.
