\chapter{Introduction}

This report presents the design, architecture, and evaluation of \textbf{M.E.R.I.T} (Multi-source Employment Ranking \& Insight Tool),
a customisable candidate screening system designed to modernise technical recruitment by moving beyond a single-source dependency.
Traditional technical recruitment is almost always dependent on the curriculum vitae (CV), which is a self-reported document, submitted by a 
candidate in a job application, in which they provide information about their education, work experience, skills, and other relevant information.
Reliance on a single source of information, especially one as unstructured and unreliable as the CV, is a major limitation of traditional
recruitment processes. MERIT aims to address this obstacle through the integration of demonstrated evidence as well as structured data 
to provide a more accurate and fair assessment of a candidate's skill and experience.

Over the past few years, there has been a rapid increase in the number of technical roles globally, leading to intensified competition
among candidates and increasing the volume of applicants that hiring managers must screen. Organisations often attempt to streamline
early-stage screening by requiring concise CV formats. While the one-page resume is standard in the US and for entry-level roles,
UK standards often allow for two pages, particularly in more experienced, technical roles. Regardless of the length requirement imposed 
in the UK and US, research often shows that CV-based screening is still demanding, even if recruiters are only given a single page. 
Due to the cognitive demands of screening, recruiters will often lean on heuristics rather than an accurate assessment, and are 
thus prone to biases \cite{moore2023, tversky1974}.

The rapid nature of manual screening often introduces decision fatigue, which leads to the amplification of the cognitive biases already
present in screening. One example of this that we can turn to is the contrast effect, in which candidate evaluation can be influenced
by the relative quality of preceding candidates \cite{palmer2014}. In the domain of technical recruitment, skills are significantly
harder to assess, and as such, these effects are further exacerbated.

Another downside of the current state of recruitment lies in the nature of the CV, particularly due to its reliability. Candidates
may inflate their responsibilities or omit weaknesses in order to present themselves as more favourable to recruiters
\cite{birkeland2006_applicant_faking_behavior}. In addition to this, there is pressure on candidates to overestimate their 
technical competencies, which is particularly suggested in programming \cite{dunning2004}. In addition to overestimations, there is 
also a lack of accountability in the natural decay of their skills over time \cite{arthur1998}, further reducing the reliability
of a manual CV-based screening process.


To address these challenges, many organisations adopt automated or semi-automated screening systems, which rank candidates based
on keywords, education, and work experience, or generate binary pass/fail outcomes. While these systems reduce recruiter workload
and accelerate the hiring process, they introduce severe limitations. They frequently ignore ``new talent signals'', such as project
portfolios, open-source contributions, or other verifiable demonstrations of ability that provide a superior assessment of technical
skills \cite{chamorro2016}. By overlooking these signals and relying on rigid scoring frameworks, automated systems systematically
underestimate qualified candidates and reinforce a narrow, one-size-fits-all approach. Another core issue is the lack of transparency 
in most systems, leaving recruiters with little or no insight into why a candidate may pass or fail. This 
black-box nature reduces trust, and also makes it hard for hiring managers to audit the fairness of automated hiring 
decisions \cite{raghavan2020}.


Automated screening platforms have been shown to inherit and propagate biases present in historical hiring data. Gender,
ethnicity, and institutional prestige can heavily influence a model's output, reproducing or even worsening the systemic discrimination present
in manual processes \cite{raghavan2020, bogen2018}. Most commercial tools also lack flexibility, applying identical scoring
frameworks to all candidates, limiting a recruiter's ability to prioritise role-specific competencies \cite{chamorro2016}.

Beyond algorithmic bias, automated screening is severely hindered by the technical fragility of the document parsing layer.
Most commercial screening systems rely on coordinate-based spatial PDF parsing libraries, such as PDFMiner \cite{pdfminer}, to
ingest unstructured CV text. These parsers frequently fail when processing modern two-column or multi-column layouts, horizontally
interleaving adjacent columns into single, corrupted strings and causing critical experience and skill details to be entirely
missed \cite{emnlp2025resumebench}. Consequently, highly qualified candidates are systematically penalised unless their CV formats 
are explicitly restricted to simple, linear designs. This layout-driven technical fragility introduces a profound, non-meritocratic 
selection bias, directly justifying the need to supplement traditional CVs with structured, verified external data sources.


The vulnerabilities to single-source evaluation, the presence of bias, and the threat of self-embellishment all point towards
the need for a multi-source, adaptable and explainable system that can assess technical skills while mitigating bias. High-validity 
selection methods have been shown to improve employee performance and productivity, leading to measurable economic and organisational 
benefits \cite{schmidt1998}. An accurate, rigorous early-stage screening process is therefore critical to reduce the likelihood of a bad hire.

MERIT addresses these issues through the combination of traditional CV data with verified behavioural and proficiency-based
evidence from multiple sources, such as GitHub and LinkedIn. By focusing on observable performance and providing transparent,
explainable justifications for its rankings, MERIT aims to reduce bias, improve reliability, and aid evidence-based
decision-making in technical recruitment.


\section{Contributions}
MERIT takes inspiration from existing screening systems commonly found in both academia and industry, notably those that map
CV information against a job description to determine candidate progression. However, rather than relying on a rigid,
single-source automated architecture or presenting a disjointed set of feature enhancements, the contribution of this work is
structured around three core architectural pillars (or processing stages). These pillars directly address the critical limitations of
contemporary automated hiring. More specifically, we address layout-driven parsing failures, single-source dependency, rigid metrics
and the lack of trust in automated rankings:

\subsubsection{Pillar 1: Extensibility \& Customisable Evaluation}
The first pillar aims to establish a dynamic, recruitment-aligned foundation by inferring screening metrics as well as target skills
directly from the job descriptions. Rather than applying a simple one-size-fits-all template, MERIT instead allows recruiters to 
rely on insights suggested by the model based on current recruitment data, however it is up to the recruiter to accept these metrics,
or to choose their own. This customisable metric extraction layer ensures flexibility within the technical hiring domain, while still
ensuring that MERIT is specific enough to capture the metrics that would be necessary for most technical professions, providing a 
highly flexible baseline for candidate ranking.

\subsubsection{Pillar 2: Multi-Source Evidence Fusion}
The second pillar of MERIT aims to address the majority of the aforementioned issues. Firstly, it attempts to overcome the 
unreliability of candidate data through the creation of a screening system that allows evidence from multiple data sources.
By integrating evidence-driven sources such as GitHub, we can evaluate candidates on demonstrated evidence, as opposed to claimed
evidence on their CV.

The next issue this pillar will attempt to address is the vulnerabilities of unstructured data parsing. Often, ATS systems
involve the parsing of CVs, however, to this day, there are significant issues with CV parsing \cite{emnlp2025resumebench}. Through the 
integration of a structured data source such as LinkedIn, we can provide supplementary data to aid a candidate's application in the 
event of malformed CV parsing.

By taking both of these factors into consideration, as well as the application of temporal skill decay and multi-source conflict
resolution, MERIT mitigates the impact of resume exaggerations as well as missing information, to help capture a more holistic
view of a candidate and their technical proficiency.

\subsubsection{Pillar 3: Glass-Box Explainability}
The final pillar revolves around trust, auditability and transparency, to combat the limitations of black-box screening systems.
MERIT abandons unexplained, arbitrary numerical grading that lacks significance in order to ensure that recruiters can truly
understand why a candidate was given a certain rank by MERIT. We adopt game-theoretic scoring (e.g.\ Shapley values) to help create this glass-box, 
pairing it with interactive recruiter dashboards, clear visualisations as well as candidate anonymisation. These additions to 
the current state of screening systems help to directly mitigate systemic bias, ensure auditability and also help to 
develop and support fair, data-driven hiring decisions.

By combining these three pillars, MERIT is able to provide a more fair and accurate assessment of a candidate's abilities
by shifting away from self-reported skill inflation and CV formatting biases. MERIT also aims to mitigate both cognitive
and discriminatory biases through candidate anonymisation with standardised formatting of CV data. These pillars will also
assist in providing complete transparency on automated screening through interactive, source-traceable recruiter reports.


\section{Report Outline}
The remainder of this report is presented as follows:
\begin{itemize}
  \item \textbf{Chapter 2: Background} outlines the standard procedures, cognitive demands, and systemic biases 
  inherent in manual recruitment.
  \item \textbf{Chapter 3: Technical Foundations \& Related Work} examines both traditional and AI-assisted automated 
  screening systems, the risks of automated bias propagation, and existing work in the field of technical recruitment that applies to MERIT.
  \item \textbf{Chapter 4: Implementation} details the architecture of MERIT, including the parsing pipeline, scoring 
  algorithms, and the dynamic metric extraction engine.
  \item \textbf{Chapter 5: Evaluation} presents the quantitative and qualitative findings validating the three core 
  architectural pillars of the system to evaluate the tradeoffs present in following MERIT's architecture.
  \item \textbf{Chapter 6: Ethics} discusses key societal, legal, and operational considerations surrounding algorithmic hiring.
  \item \textbf{Chapter 7: Critical Reflection \& Future Work} provides a pillar-aligned retrospective on evaluation
  outcomes, design trade-offs, and methodological scope, together with a prioritised research and deployment roadmap.
  \item \textbf{Chapter 8: Conclusion} delivers a final verdict on whether MERIT achieved its three pillars
  (extensible, multi-source, explainable) as an auditable screening prototype.
\end{itemize}